---
title: The genetics of behavioral flexibility across the range of a rapidly expanding species
author: '[Dr. Corina Logan](http://CorinaLogan.com) (Max Planck Institute for Evolutionary Anthropology, corina_logan@eva.mpg.de), Carolyn Rowney (University of California Santa Barbara / Max Planck Institute for Evolutionary Anthropology), Dr. Zoe Johnson-Ulrich (University of California Santa Barbara / Max Planck Institute for Evolutionary Anthropology), Luisa Bergeron (University of California Santa Barbara / Max Planck Institute for Evolutionary Anthropology), Dr. Dieter Lukas (Max Planck Institute for Evolutionary Anthropology)'
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
  word_document: default
bibliography: /Users/corina/GTGR/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

###A. STATE OF THE DATA

This preregistration was written prior to collecting any data.

###B. PARTITIONING THE RESULTS

We may publish the results from each hypothesis separately.

###C. HYPOTHESES

###D. METHODS

###E. ANALYSIS PLAN

###F. PLANNED SAMPLE

Great-tailed grackles (n > 200) will be caught in the wild at three field sites across their geographic range: the center of their original range (Central America), the middle of the northward expanding edge (Tempe, Arizona USA), and on the northern expanding edge (to be determined). Individuals will be identified using colored leg bands in unique combinations, their data collected (blood, feathers, and biometrics), and then they will be released back to the wild. Some individuals (40-100) will be brought temporarily into aviaries for behavioral testing, and then they will be released back to the wild. 

**Sample size rationale**

We will band as many birds as possible throughout the year, and conduct behavioral tests in aviaries (during the non-breeding seasons) and in the wild (year-round) on as many birds as we can at each field site. The minimum sample size will be 200 banded birds and 40 behavior-tested birds in total, however we expect to be able to sample many more.

**Data collection stopping rule**

We will stop collecting data in April 2023 when the current funding ends, or after data have been collected at all three field sites, whichever date comes first.

###G. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017] and SP606267 [2018])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)
5) University of Cambridge ethical review process (non-regulated use of animals in scientific procedures: zoo4/17)

###H. AUTHOR CONTRIBUTIONS

**Logan:** Hypothesis development, study design, materials, data collection, data analysis and interpretation, write up, funding.

**Rowney:** Data collection, sample processing, data analysis and interpretation, editing/revising.

**Johnson-Ulrich:** Data collection, sample processing, data analysis and interpretation, editing/revising.

**Bergeron:** Data collection, sample processing, data analysis and interpretation, editing/revising.

**Lukas:** Hypothesis development, study design, data analysis and interpretation, write up.

###I. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

###J. ACKNOWLEDGEMENTS

We thank Ben Trumble for hosting the grackle project at Arizona State University (providing office and lab space); Melissa Wilson Sayers for sponsoring our affiliations at Arizona State University and lending lab equipment; Arizona State University School of Life Sciences Department Animal Care and Technologies for providing space for our aviaries and for their excellent support of our daily activities; and Cornell University (Irby Lovette and Bronwyn Butcher) for letting us use their lab to process the DNA and for teaching us how.

###K. REFERENCES
