---
title: "Are the more flexible individuals also better at inhibition?"
author: "[Dr. Corina Logan](http://CorinaLogan.com) (University of Cambridge, cl417@cam.ac.uk) and [Dr. Claudia Wascher](https://www.claudiawascher.com) (Anglia Ruskin University)"
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
  word_document: default
bibliography: /Users/corina/R/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

###A. STATE OF THE DATA

This preregistration was written prior to collecting any data.

###B. PARTITIONING THE RESULTS

We may present these results separately: 1) determining whether grackles have inhibition abilities, and 2) linking variation in performance on inhibition tasks to measures of flexibility.

###C. HYPOTHESIS

**If flexibility requires inhibition, then individuals that are more behaviorally flexible, as measured by reversal learning and switching between options on a multi-access box, will also be better at inhibiting their responses in three tasks: delayed gratification task, go no-go task, and detour task.**

**P1:** Individuals that are faster to reverse preferences on a reversal learning task and who also have lower latencies to successfully solve new loci after previously solved loci become unavailable (multi-access box) (see [flexibility preregistration](./g_flexmanip.Rmd)) will perform better in the go no-go task (methods similar to @harding2004animal), in the detour task (methods as in @maclean2014evolution who call it the "cylinder task"), and they will wait longer for higher quality (more preferred) food, but not for higher quantities (methods as in @hillemann2014waiting). Waiting for higher quality food has been validated as a test of inhibition in birds, while waiting for a higher quantity of food does not appear to measure inhibition (@hillemann2014waiting).

**P1 alternative 1:** If there is no correlation between flexibility and performance on the inhibition tasks, this may indicate that the flexibility tasks may not require much inhibition (particularly if the inhibition results are reliable - see *P1 alternative 2*).

**P1 alternative 2:** If there is no correlation between flexibility and performance on the inhibition tasks, this may indicate that the inhibition tasks had low reliability and were therefore too noisy to correlate with flexibility.

**P2:** If there is no correlation in performance across inhibition tasks, it may indicate that that one or more of these tasks does not measure inhibition, or that they measure different types of inhibition (see @friedman2004relations).

**P2 alternative:** If go no-go task performance strongly correlates with performance on the delayed gratification task, this indicates these two tasks measure the same trait, which therefore validates an inhibition task using a touch screen (the go no-go task).

**P3:** If individuals perform well on the detour task and with little individual variation, this is potentially because they will have had extensive experience looking into the sides of opaque tubes during reversal learning. To determine whether prior experience with opaque tubes in reversal learning contributed to their detour performance, a subset of individuals will experience the detour task before any reversal learning tests. If this subset performs the same as the others, then previous experience with tubes does not influence detour task performance. If the subset performs worse than the others, this indicates that detour task performance depends on the previous experiences of the indviduals tested.

###D. METHODS

####**Randomization and counterbalancing** 

**P3** 

Two individuals from each batch will experience the detour task before participating in the flexibility manipulation. These individuals will be randomly selected using the random number generator at https://www.random.org.

**P1-P2**

For the rest of the individuals (n=6 per batch), the order of the three inhibition tasks will be counterbalanced across birds (using https://www.random.org to randomly assign individuals to one of three experimental orders). 1/3 of the individuals will experience:

1. Delayed gratification task

2. Go no-go task

3. Detour

1/3 of the individuals will experience:

1. Go no-go task

2. Detour

3. Delayed gratification task

1/3 of the individuals will experience: 

1. Detour

2. Delayed gratification task

3. Go no-go task

**Delayed gratification** 

- Food preference test: food will be presented in random combinations over six sessions of 12-15 trials. 

- Training trials: The type of demonstration and training trials varied randomly (with more demo trials near the beginning of training), incorporating trials in which food of the same sort accumulated (quantity), food of ascending quality accumulated (quality), and trials in which we added increasingly larger food pieces throughout the trial (size)

- Test: we will test each food quality (low, mid, high) twice in randomized order in each session.

**Go no-go** 

Go and no-go trials will be presented randomly with the restriction that no more than four of the same type will occur in a row. The rewarded color will be counterbalanced across birds.

**Detour** 

The side from which the apparatus is baited will be consistent within subjects, but counterbalanced across subjects.

####**Blinding of conditions during analysis**

No blinding is involved in this study.

####**Dependent variables**

#####*P1: the more flexible individuals are better at inhibition*

1) **Delayed gratification:** Number of food pieces waited for (0-3). A successful wait is defined as waiting for at least one additional piece of food to be added to the serving lid of the three possible additional food items, and accepted at least one piece of the reward pieces.

2) **Go no-go: ** 

  a) The number of trials to reach criterion (85% correct) where correct responses involve pecking when the rewarded stimulus is displayed and not pecking when the unrewarded stimulus is displayed, and incorrect responses involve pecking when the unrewarded stimulus is displayed, and not pecking when the rewarded stimulus is displayed

  b) The latency to respond (peck the target key)

3) **Detour:** First approach (physical contact): Correct (to the tube's side opening) or Incorrect (to the front unopen area of the tube) (methods as in @maclean2014evolution).

One model will be run per dependent variable.

#####*P3: does training improve detour performance?*

1) First approach (physical contact): Correct (to the tube's side opening) or Incorrect (to the front unopen area of the tube) (methods as in @maclean2014evolution).

####**Independent variables**

#####*P1: delayed gratification*

1) Food quality or quantity (Quality: High, Med, Low; Quantity: Smaller, Medium, Larger)

2) Trial

3) Delay (2, 5, 10, 20, 40, 60, or 80 seconds)

4) Flexibility 1: **Number of trials to reverse** a preference in the last reversal an individual experienced (reversal learning; an individual is considered to have a preference if it chose the rewarded option at least 17 out of the most recent 20 trials, with a minimum of 8 or 9 correct choices out of 10 on the two most recent sets of 10 trials). See behavioral flexibility [preregistration](./g_flexmanip.Rmd).

5) Flexibility 2: The **ratio of correct divided by incorrect trials** for the first 40 trials in their final reversal after the individual has seen the newly rewarded option once. These 40 trials include trials where individuals were offered the test and chose not to participate (i.e., make a choice). This accounts for flexibility that can occur when some individuals inhibit their previously rewarded preference (thus exhibiting flexibility because they changed their behavior when circumstances changed), but are not as exploratory as those who have fewer 'no choice' trials. 'No choice' data is data that is otherwise excluded from standard reversal learning analyses. Including 'no choice' trials, controls for individual differences in exploration because those that refuse to choose are not exploring new options, which would allow them to learn the new food location.

6) Flexibility 3: If the number of trials to reverse a preference does not positively correlate with the latency to attempt or solve new loci on the multi-access box (an additional measure of behavioral flexibility), then the **average latency to solve** and the **average latency to attempt** a new option on the multi-access box will be additional dependent variables. See behavioral flexibility [preregistration](./g_flexmanip.Rmd).

#####*P1: go no-go*

Model 2a: number of trials to reach criterion

1) Flexibility 1: Number of trials to reverse a preference in the last reversal an individual experienced (reversal learning; as above)

2) Flexibility 2: The **ratio of correct divided by incorrect trials** for the first 40 trials in their final reversal after the individual has seen the newly rewarded option once (reversal learning; as above).

3) Flexibility 3: If the number of trials to reverse a preference does not positively correlate with the latency to attempt or solve new loci on the multi-access box, then the **average latency to solve** and the **average latency to attempt** a new option on the multi-access box will be additional independent variables (as above).

Model 2b: latency to respond

1) Correct or incorrect response

2) Trial

3) [Flexibility Condition](./g_flexmanip.Rmd): control, manipulated

3) ID (random effect because multiple measures per bird)

#####*P1: detour*

1) Trial

2) Flexibility 1: Number of trials to reverse a preference in the last reversal an individual experienced (reversal learning; as above)

3) Flexibility 2: The **ratio of correct divided by incorrect trials** for the first 40 trials in their final reversal after the individual has seen the newly rewarded option once (reversal learning; as above).

4) Flexibility 3: If the number of trials to reverse a preference does not positively correlate with the latency to attempt or solve new loci on the multi-access box, then the **average latency to solve** and the **average latency to attempt** a new option on the multi-access box will be additional independent variables (as above).

#####*P3: does training improve detour performance?*

1) Condition: pre- or post-reversal learning tests

###E. ANALYSIS PLAN

We do not plan to **exclude** any data. When **missing data** occur, the existing data for that individual will be included in the analyses for the tests they completed. Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam). 

####*Data checking*

The data will be visually checked to determine whether they are normally distributed via two methods: 1) normality is indicated when the histograms of actual data match those with simulated data (Figure 2), and 2) normality is indicated when the residuals closely fit the dotted line in the Normal Q-Q plot (Figure 3) [@zuur2009].

```{r dist_checkHist, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, fig.cap="**Figure 2.** Part 1. The distribution of the actual data versus what a normal distribution would look like with simulated data. Part 2. Residuals vs fitted: checking for homogeneity, which is satisfied if residuals have an even spread across the x-axis; Normal Q-Q: residuals are normally distributed if they are on the diagonal line; Residuals vs leverage: Cook's distance <1 means no influential observations [@zuur2009]."}
acc <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_accumulation.csv", header=T, sep=",", stringsAsFactors=F) 

go <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_go.csv", header=T, sep=",", stringsAsFactors=F) 

detour <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_detour.csv", header=T, sep=",", stringsAsFactors=F) 


#Check the dependent variables for normality: Histograms
op <- par(mfrow=c(2,3), mar=c(4,4,2,0.2))
#This is what the distribution of actual data looks like
hist(acc$NumberOfAccumulationsWaited, xlab="Delay: Number of accumulations waited", main="Actual Data")
hist(go$CorrectResponse, xlab="Go/no-go: Number of correct responses", main="Actual Data")
hist(detour$FirstApproach, xlab="Detour: First approach", main="Actual Data")

#Given the actual data, this is what a normal distribution would look like
X2 <- rnorm(1281, mean=mean(acc$NumberOfAccumulationsWaited), sd=sd(acc$NumberOfAccumulationsWaited))
hist(X2, xlab="Delay: No. accumulations waited", main="Simulated Data")

Y2 <- rnorm(1281, mean=mean(go$CorrectResponse), sd=sd(go$CorrectResponse))
hist(Y2, xlab="Go/no-go: No. correct responses", main="Simulated Data")

Z2 <- rnorm(1281, mean=mean(detour$FirstApproach), sd=sd(detour$FirstApproach))
hist(Z2, xlab="Detour: First approach", main="Simulated Data")


#Check the dependent variables for normality: Q-Q plots
op <- par(mfrow=c(3,4), mar=c(4,4,2,0.2))
plot(glm(acc$NumberOfAccumulationsWaited~acc$Delay)) 
plot(glm(go$CorrectResponse~go$Trial)) 
plot(glm(detour$FirstApproach~detour$Trial)) 
```

If the data do not appear normally distributed, visually check the residuals. If they are patternless, then assume a normal distribution (Figure 4) [@zuur2009].

```{r dist_checkResid, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, fig.cap="**Figure 4.** Visual check of the residuals."}
#Check the dependent variables for normality: Residuals
detour <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_detour.csv", header=T, sep=",", stringsAsFactors=F) 

acc <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_accumulation.csv", header=T, sep=",", stringsAsFactors=F) 

go <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_go.csv", header=T, sep=",", stringsAsFactors=F) 

#Figure 3. Visual check of the residuals
op <- par(mfrow=c(1,3), mar=c(4,4,2,0.2))
plot(residuals(glm(detour$FirstApproach~detour$Trial)), ylab="Detour residuals: First approach ~ Trial")
plot(residuals(glm(acc$NumberOfAccumulationsWaited~acc$Delay)), ylab="Delay residuals: Number of accumulations waited ~ Delay")
plot(residuals(glm(go$CorrectResponse~go$Trial)), ylab="Go/no-go: Residuals Correct response ~ Trial")
```

####*P1: delayed gratification*

**Assess food preferences:** Conduct preference tests between pairs of different foods. Rank food preferences into three categories (High, Medium, Low) in the order of the percentage of times a food was chosen.

**Analysis:** Generalized Linear Model (GLM; glm function, stats package) with a Poisson distribution and log link. The contribution of each independent variable will be evaluated using the Estimate in the full model.

```{r better, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
acc <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_accumulation.csv", header=T, sep=",", stringsAsFactors=F) 

#GLM
better <- glm(NumberOfAccumulationsWaited ~ Delay + FoodQualityQuantity + Trial + TrialsToReverseLast + FlexRatio, family="poisson", data=acc)
#summary(better)

better1 <- summary(better)
library(xtable)
better1.table <- xtable(better1)
library(knitr)
kable(better1.table, caption="Table U: Model selection output.", format="html", digits=2)
```

####*P1: go no-go*

**Analysis:** 

**Model 2a:** Generalized Linear Model (GLM; glm function, stats package) with a Poisson distribution and a log link. The contribution of each independent variable will be evaluated using the Estimate in the full model.

```{r go, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
go <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_go.csv", header=T, sep=",", stringsAsFactors=F) 

#GLM
go1 <- glm(TrialsToCriterion ~ TrialsToReverseLast + FlexRatio, family="poisson", data=go)
sgo1 <- summary(go1)
library(xtable)
sgo1.table <- xtable(sgo1)
library(knitr)
kable(sgo1.table, caption="Table T: Model selection output.", format="html", digits=2)
```

**Model 2b:** A Generalized Linear Mixed Model (GLMM; MCMCglmm function, MCMCglmm package; [@hadfieldMCMCglmmpackage]) will be used with a Poisson distribution and log link using 13,000 iterations with a thinning interval of 10, a burnin of 3,000, and minimal priors (V=1, nu=0) [@hadfield2014coursenotes]. I will ensure the GLMM shows acceptable convergence (lag time autocorrelation values <0.01; [@hadfieldMCMCglmmpackage]), and adjust parameters if necessary. The contribution of each independent variable will be evaluated using the Estimate in the full model.

```{r golatency, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
go <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_golatency.csv", header=T, sep=",", stringsAsFactors=F) 

#GLM
library(MCMCglmm)
prior = list(R=list(R1=list(V=1,nu=0),R2=list(V=1,nu=0),R3=list(V=1,nu=0)), G=list(G1=list(V=1,nu=0))) 

golat <- MCMCglmm(LatencyToRespond ~ CorrectResponse * Trial * FlexibilityCondition, random=~ID, family="poisson", data=go, verbose=F, prior=prior, nitt=13000, thin=10, burnin=3000)
summary(golat)
autocorr(golat$Sol) #Did fixed effects converge?
autocorr(golat$VCV) #Did random effects converge?
```

####*P1: detour*

**Analysis:** Generalized Linear Model (GLM; glm function, stats package) with a binomial distribution and a logit link. The contribution of each independent variable will be evaluated using the Estimate in the full model.

```{r detour, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
detour <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_detour.csv", header=T, sep=",", stringsAsFactors=F) 

#GLM
detour$ID <- factor(detour$ID)
de <- glm(FirstApproach ~ Trial + TrialsToReverseLast + FlexRatio, family="binomial", data=detour)
sde <- summary(de)
library(xtable)
sde.table <- xtable(sde)
library(knitr)
kable(sde.table, caption="Table T: Model selection output.", format="html", digits=2)
```

#####*P1 alternative 2: are inhibition results reliable?*

The reliability of the inhibition tests will be calculated using Cronbach's Alpha (as in @friedman2004relations; R package: psych [@psych], function: alpha), which is indicated by std.alpha in the output.

```{r reliability, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
rel <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_inhibition.csv", header=T, sep=",", stringsAsFactors=F) 

library(psych)
reliab <- alpha(rel, check.keys=TRUE) #Check.keys automatically reverses the coding for variables that are negatively correlated with the total scale. How to interpret: http://data.library.virginia.edu/using-and-interpreting-cronbachs-alpha/ "If all of the scale items are entirely independent from one another (i.e., are not correlated or share no covariance), then alpha = 0; and, if all of the items have high covariances, then alpha will approach 1 as the number of items in the scale approaches infinity. In other words, the higher the alpha coefficient, the more the items have shared covariance and probably measure the same underlying concept."
summary(reliab)
#Insert into text: `r reliab$std.alpha`
```

When comparing all three tests, alpha= .

####*P2: correlation across inhibition tasks*

**See analysis description for P1 alternative 2.** 

```{r reliability2, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
rel2 <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_inhibition2.csv", header=T, sep=",", stringsAsFactors=F) 

library(psych)
reliab2 <- alpha(rel2, check.keys=TRUE)
summary(reliab2)
#Insert into text: `r reliab2$std.alpha`
```

When analyzing only the delayed gratification and go no-go tasks, the reliability is alpha= .

#####*P3: does training improve detour performance?*

**Analysis:** Generalized Linear Model (GLM; glm function, stats package) with a binomial distribution and a logit link. The contribution of each independent variable will be evaluated using the Estimate in the full model.

```{r detour2, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
detour <- read.csv ("/Users/corina/R/GrackleTestsYear1/Cognition/data_detour.csv", header=T, sep=",", stringsAsFactors=F) 

#GLM
de <- glm(FirstApproach ~ Condition, family="binomial", data=detour)
sde <- summary(de)
library(xtable)
sde.table <- xtable(sde)
library(knitr)
kable(sde.table, caption="Table T: Model selection output.", format="html", digits=2)
```

####*Alternative Analyses*

We anticipate that we will want to run additional/different analyses after reading @statrethinkingbook. We will revise this preregistration to include these new analyses before conducting the analyses above.

###F. PLANNED SAMPLE

Great-tailed grackles will be caught in the wild in Tempe, Arizona USA for individual identification (colored leg bands in unique combinations). Some individuals (16-40) will be brought temporarily into aviaries for testing, and then they will be released back to the wild.

**Sample size rationale**

We will test as many birds as we can in the two years we have at this field site given that the birds only participate in tests in aviaries during the non-breeding season (September through February). The minimum sample size will be 16, however we expect to be able to test many more.

**Data collection stopping rule**

We will stop testing birds in March 2019.

###G. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017] and SP606267 [2018])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)
5) University of Cambridge ethical review process (non-regulated use of animals in scientific procedures: zoo4/17)

###H. AUTHOR CONTRIBUTIONS

**Logan:** Hypothesis development, experimental design (go no-go task), data collection, data analysis and interpretation, write up, materials/funding.

**Wascher:** Hypothesis development, experimental design (delayed gratification and detour tasks), data analysis and interpretation, write up.

###I. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology, and by a Leverhulme Early Career Research Fellowship to Logan.

###J. ACKNOWLEDGEMENTS

We thank Dieter Lukas for help polishing the predictions; Jay Taylor for grackle scouting, hosting the grackle project at Arizona State University, providing office and lab space, and co-supervising students; and Arizona State University School of Life Sciences Department Animal Care and Technologies for providing space for our aviaries and for their excellent support of our daily activities.

###J. REFERENCES