---
title: Validating morphological condition indices and their relationship with reproductive success in great-tailed grackles
author: 'Berens JM (Arizona State University), [Logan CJ](http://CorinaLogan.com) (Max Planck Institute for Evolutionary Anthropology, MPI EVA, corina_logan@eva.mpg.de), Folsom M (MPI EVA), Bergeron L (University of California Santa Barbara, UCSB), McCune KB (UCSB, kelseybmccune@gmail.com)'
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    code_folding: hide 
  md_document: 
    toc: true
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  github_document: 
    toc: true
bibliography: /Users/corina/GitHub/grackles/Files/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

***Click [here](https://github.com/corinalogan/grackles/blob/master/README.md) to navigate to the version-tracked reproducible manuscript (.Rmd file)***

### ABSTRACT
Morphological variation among individuals has the potential to influence multiple life history characteristics such as dispersal, migration, reproductive fitness, and survival (@wilder2016moving). Theoretically, individuals that are in better “condition” (i.e. fat reserves, @labocha2012morphometric) should be able to disperse or migrate further or more successfully, have greater reproductive fitness, and survive for longer (@wilder2016moving). Researchers have used a variety of morphological proxy variables to quantify condition (i.e., fat score, weight, ratio of weight to tarsus length, ratio of weight to wing chord length, @labocha2014which), however, there is mixed support regarding whether these proxy variables relate to life history characteristics (@wilder2016moving; @labocha2014which). Additionally, although some researchers use multiple morphological proxy variables for condition (i.e. @warnock1998spring), rarely has there been direct comparisons among proxies to validate that they measure the same trait. In this investigation, we will compare two condition proxies (fat score and the ratio of weight to tarsus length) to validate whether they measure the same trait in our study system, the great-tailed grackle (Quiscalus mexicanus). We will then test whether our morphological proxy variables correlate with reproductive success, measured as whether a female had a fledgling or not and whether a male held a territory containing nests or not. Results will improve our understanding of measures of condition in grackles, and birds in general, and the importance of condition for reproductive success - a necessary component for selection to act.

### A. STATE OF THE DATA

This preregistration uses secondary data that are collected as part of other ongoing  investigations (tarsus length in [flex genes](http://corinalogan.com/Preregistrations/g_flexgenes.html); tarsus length, body weight, number of fledglings, and whether a male holds a territory in [within population](http://corinalogan.com/Preregistrations/g_withinpop.html); and tarsus length in [expansion](http://corinalogan.com/Preregistrations/g_expansion.html)). This preregistration was written (July 2019) and submitted to Peer Community In Ecology for pre-study peer review (August 2019) before any analyses were conducted.

### B. HYPOTHESES

#### H1: There is a relationship between two different morphological indices of condition: fat score and the ratio of body weight to tarsus length.

**Prediction 1:**  Fat score and the ratio of weight to tarsus length will be positively correlated. This would indicate that these two indices measure the same trait, and it is likely they both are proxies for fat content.

**Prediction 1 alternative 1:** There is a negative correlation between fat score and the ratio of weight to tarsus length. This would indicate that there may be a tradeoff between the two indices where a higher weight to tarsus length ratio may measure muscle content rather than fat, and individuals with more muscle have less visible fat.

**Prediction 1 alternative 2:** There is no correlation between fat score and the ratio of weight to tarsus length. This indicates that these two variables do not measure the same trait. Fat score may not adequately capture a bird’s condition because birds may be selected to only store the minimal fat necessary to prevent starvation, while also minimizing the weight gain that would make them easier targets for predators (@barnett2015mass). Similarly, the ratio of weight to tarsus length has been shown to be heavily influenced by body size, therefore reflecting structural size rather than fat storage (@labocha2012morphometric).

#### H2: Condition (as measured by fat score and the ratio of weight to tarsus length) relates to reproductive success (measured as a binary variable of whether a female had one or more fledglings (1) or not (0), and whether a male defended a territory containing nests (1) or not (0)).

**Prediction 2:** Morphological indices of condition (fat score and the ratio of weight to tarsus length) will correlate positively with reproductive success. This would indicate that individuals with more fat, and therefore higher energy reserves, are better able to acquire the resources necessary for reproduction.

**Prediction 2 alternative 1:** Morphological indices of condition (fat score and the ratio of weight to tarsus length) will correlate negatively with reproductive success. This indicates that individuals may make trade offs, with some acquiring more food and increasing their energy reserves, and others prioritizing reproductive activities over increasing energy reserves.

**Prediction 2 alternative 2:** Morphological indices of condition (fat score and the ratio of weight to tarsus length) do not correlate with reproductive success. This indicates that other, potentially non-morphological, individual characteristics relate to reproductive success (i.e., cognition, nest site selection, breeding experience, predator vigilance, etc.).

### C. METHODS

#### **Planned Sample**

Great-tailed grackles are caught in the wild in Tempe, Arizona using a variety of methods (e.g., walk-in trap, bownet, mist net). After capture we immediately process birds by attaching colored leg bands in unique combinations for individual identification, conducting morphological measurements of weight, tarsus and wing chord length; tail, skull, and bill length and fat score (the amount of visible fat under the skin in the clavicle and abdomen as in @kaiser1993new). Most grackles are released after completion of color band marking, measurements, and acquiring a blood sample. A subset of grackles are held in aviaries for up to 6 months for behavioral testing, and then released back to the wild at their location of capture.

From March - August, we monitor the behavior of all color-marked grackles to determine their nesting status. We follow females carrying nesting materials to find their nest, then we check each nest approximately every day to determine the status (building, incubation, nestlings, fledglings, failed).

Individuals included in this sample will be those for which we have measures of condition when they were adults. We will not include individuals whose data were collected as juveniles. As of 30 July 2019, we have fledgling data for 14 females that exhibited breeding behavior (5 had 1+ fledgling, 9 had no fledglings) and breeding territory status for 9 males (7 territory holders, 3 non-territory holders, 2 not observed so not part of this sample).

#### **Sample size rationale**

We will continue to color mark as many grackles as possible, and collect biometric data and fat scores. The minimum sample size for H2 will be 57 marked grackles because that is how many individuals we have biometric and breeding behavior data for so far. However, we expect to be able to catch many more (20-50) before the end of this investigation in Tempe, Arizona that we can use in H1.

#### **Data collection stopping rule**

We will stop collecting data for this project in early August 2021, at the conclusion of the second breeding season in Tempe, Arizona.

#### **Open materials**

Biometric measurement protocol:
https://gitlab.com/corinalogan/the-grackle-project/blob/master/protocolBiometrics.pdf 

Nest check protocol:
https://gitlab.com/corinalogan/the-grackle-project/blob/master/protocolNestCheck.pdf 

####**Open data**

When the study is complete, the data will be published in the Knowledge Network for Biocomplexity's data repository.

#### **Randomization and counterbalancing**

There is no randomization or counterbalancing in this investigation. 

#### **Blinding of conditions during analysis**

No blinding is involved in this investigation.

#### **Dependent Variables**

**P1: correlation between fat and the ratio of weight to tarsus length**

1) Fat score (the amount of visible fat under the skin in the clavicle and abdomen reported as a score from 0 (no fat) to 8 (fat completely covers muscles and underside of the bird); @kaiser1993new)

**P2: condition and reproductive success**

1) Female had one or more fledglings (yes, no)

2) Male held territory containing nests (yes, no)


#### **Independent Variables**

**P1: correlation between fat and the ratio of body weight to tarsus length**

1) Ratio of body weight to tarsus length (tarsus length = average of left and right legs as in @bleeker2005body)

2) Season (breeding [Sep-Feb], non-breeding [Mar-Aug])

3) Temporarily held in aviaries for behavioral testing at any point during this study (yes, no)

4) Random effect: Experimenter (because several different experimenters measure dependent variables on multiple different birds)

**P2: condition and reproductive success**

1) Fat score (note: if the fat score and the ratio of body weight to tarsus length are positively correlated, then we will use only fat score in the model for P2. If they are not positively correlated, then we will add the ratio as an independent variable in the P2 model)

2) Temporarily held in aviaries for behavioral testing at any point during this study (yes, no)

3) Random effect: Year (to determine whether conditions in a given breeding season similarly affected all grackle behavior and nest success)

4) Random effect: Bird ID (because there will be multiple measures of reproductive success for each bird)

### D. ANALYSIS PLAN

We will **exclude** data that was collected from the grackles when they were released from the aviaries to avoid any confounds due to their time in the aviary (e.g., perhaps unlimited nutritious food in the aviaries decreased their fat score). Therefore, all data included in this study will come only from wild-caught grackles (including the birds that were brought into the aviaries on their first capture). When **missing data** occur, the existing data for that individual will be included in the analyses for which their data exist. Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam). When there is more than one experimenter collecting data, experimenter will be added as a random effect to account for potential differences between experimenters in conducting the sessions.

####*Data checking*

The data will be checked for overdispersion, underdispersion, zero-inflation, and heteroscedasticity with the DHARMa R package [@Hartig2019dharma] following methods by [Hartig](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html). 

#### *P1 analysis: correlation between fat and the ratio of weight to tarsus length*

**Analysis:** We use a Generalized Linear Mixed Model (GLMM; MCMCglmm function, MCMCglmm package; [@hadfield2010mcmc]) with a binomial distribution (called “categorical” in MCMCglmm) and log link using 130,000 iterations with a thinning interval of 10, a burnin of 30,000, and minimal priors (V=1, nu=0) [@hadfield2014coursenotes]. We will ensure the GLMM shows acceptable convergence (lag time autocorrelation values <0.01; [@hadfield2010mcmc]), and adjust parameters if necessary to meet this criterion. We will determine whether an independent variable had an effect or not using the Estimate in the full model.

```{r p1, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
d <- read.csv ("/Users/corina/GTGR/data/data_condition.csv", header=T, sep=",", stringsAsFactors=F) 

#Factor the random effect variables
Experimenter <- as.factor(d$Experimenter)

# DATA CHECKING
library(DHARMa)
library(lme4)
library(MCMCglmm)

prior = list(R=list(R1=list(V=1,nu=0),R2=list(V=1,nu=0)), G=list(G1=list(V=1,nu=0)))

#Data checking for GLMM
simulationOutput <- simulateResiduals(fittedModel = MCMCglmm(FatScore ~ Body + Season + Aviary, random=~Experimenter, family="poisson", data=d, verbose=F, prior=prior, nitt=130000, thin=10, burnin=30000), n=250) #250 simulations, but if want higher precision change n>1000
simulationOutput$scaledResiduals #Expect a flat distribution of the overall residuals, and uniformity in y direction if plotted against any predictor
testDispersion(simulationOutput) #if under- or over-dispersed, then p-value<0.05, but then check the dispersion parameter and try to determine what in the model could be the cause and address it there, also check for zero inflation
testZeroInflation(simulationOutput) #compare expected vs observed zeros, not zero-inflated if p<0.05
testUniformity(simulationOutput) #check for heteroscedasticity ("a systematic dependency of the dispersion / variance on another variable in the model" Hartig, https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html), which is indicated if dots aren't on the red line and p<0.05. Also...
plot(simulationOutput) #...there should be no pattern in the data points in the right panel
plotResiduals(Body, simulationOutput$scaledResiduals) #plot the residuals against other predictors (in cases when there is more than 1 fixed effect) - can't get this code to work yet
plotResiduals(Season, simulationOutput$scaledResiduals) #plot the residuals against other predictors (in cases when there is more than 1 fixed effect) - can't get this code to work yet

#GLMM 
m <- MCMCglmm(FatScore ~ Body + Season + Aviary, random=~Experimenter, family="poisson", data=d, verbose=F, prior=prior, nitt=130000, thin=10, burnin=30000)
summary(m)
#autocorr(m$Sol) #Did fixed effects converge?
#autocorr(m$VCV) #Did random effects converge?
```


#### *P2 analysis: condition and repro success*

A GLMM is conducted as in P1, except this GLMM used a binomial distribution (called "categorical" in MCMCglmm) due to the response variables being categorized as only yes or no. If fat score and the ratio of body weight to tarsus length are positively correlated in P1, then we will use only fat score as the independent variable in this GLMM. If they are not positively correlated, we will include both as independent variables.

```{r p2, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
d <- read.csv ("/Users/corina/GTGR/data/data_condition.csv", header=T, sep=",", stringsAsFactors=F) 

#Factor the random effect variables
Year <- as.factor(d$Year)
ID <- as.factor(d$ID)

# DATA CHECKING
library(DHARMa)
library(lme4)
library(MCMCglmm)

prior = list(R=list(R1=list(V=1,nu=0),R2=list(V=1,nu=0)), G=list(G1=list(V=1,nu=0)))

#Data checking for GLMM: fledglings 
simulationOutput <- simulateResiduals(fittedModel = MCMCglmm(Fledglings ~ FatScore + Aviary, random=~Year + ID, family="categorical", data=d, verbose=F, prior=prior, nitt=130000, thin=10, burnin=30000), n=250) #250 simulations, but if want higher precision change n>1000
simulationOutput$scaledResiduals #Expect a flat distribution of the overall residuals, and uniformity in y direction if plotted against any predictor
testDispersion(simulationOutput) #if under- or over-dispersed, then p-value<0.05, but then check the dispersion parameter and try to determine what in the model could be the cause and address it there, also check for zero inflation
testZeroInflation(simulationOutput) #compare expected vs observed zeros, not zero-inflated if p<0.05
testUniformity(simulationOutput) #check for heteroscedasticity ("a systematic dependency of the dispersion / variance on another variable in the model" Hartig, https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html), which is indicated if dots aren't on the red line and p<0.05. Also...
plot(simulationOutput) #...there should be no pattern in the data points in the right panel

#Data checking for GLMM: territory holder
simulationOutput <- simulateResiduals(fittedModel = MCMCglmm(Territory ~ FatScore + Aviary, random=~Year + ID, family="categorical", data=d, verbose=F, prior=prior, nitt=130000, thin=10, burnin=30000), n=250) #250 simulations, but if want higher precision change n>1000
simulationOutput$scaledResiduals #Expect a flat distribution of the overall residuals, and uniformity in y direction if plotted against any predictor
testDispersion(simulationOutput) #if under- or over-dispersed, then p-value<0.05, but then check the dispersion parameter and try to determine what in the model could be the cause and address it there, also check for zero inflation
testZeroInflation(simulationOutput) #compare expected vs observed zeros, not zero-inflated if p<0.05
testUniformity(simulationOutput) #check for heteroscedasticity ("a systematic dependency of the dispersion / variance on another variable in the model" Hartig, https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html), which is indicated if dots aren't on the red line and p<0.05. Also...
plot(simulationOutput) #...there should be no pattern in the data points in the right panel

#GLMM fledglings
m1 <- MCMCglmm(Fledglings ~ FatScore + Aviary, random=~Year + ID, family="categorical", data=d, verbose=F, prior=prior, nitt=130000, thin=10, burnin=30000)
summary(m1)
#autocorr(m1$Sol) #Did fixed effects converge?
#autocorr(m1$VCV) #Did random effects converge?

#GLMM territory holder
m2 <- MCMCglmm(Territory ~ FatScore + Aviary, random=~Year + ID, family="categorical", data=d, verbose=F, prior=prior, nitt=130000, thin=10, burnin=30000)
summary(m2)
#autocorr(m2$Sol) #Did fixed effects converge?
#autocorr(m2$VCV) #Did random effects converge?
```

###E. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017], SP606267 [2018], and SP639866 [2019])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)

###F. AUTHOR CONTRIBUTIONS

**Berens:** Data collection, revising/editing.

**Logan:** Study design, write up, revising/editing, materials/funding.

**Folsom:** Data collection, revising/editing.

**Bergeron:** Data collection, revising/editing.

**McCune:** Hypothesis development, data collection, write up, revising/editing.

###G. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

###I. ACKNOWLEDGEMENTS

We thank our research assistants for help with trapping the grackles and collecting the biometric and nest/territory data: Aelin Mayer, Nancy Rodriguez, Brianna Thomas, Aldora Messinger, Elysia Mamola, Michael Guillen, Rita Barakat, Adriana Boderash, Olateju Ojekunle, August Sevchik, Justin Huynh, Amanda Overholt, and Michael Pickett.

###J. [REFERENCES](MyLibrary.bib)
