---
title: Investigating sex differences in genetic diversity in great-tailed grackles in Tempe, Arizona to infer potential sex biases in dispersal
author: 'August Sevchik (Arizona State University), [Dr. Corina Logan](http://CorinaLogan.com) (Max Planck Institute for Evolutionary Anthropology, MPI EVA, corina_logan@eva.mpg.de), [Dr. Dieter Lukas](http://dieterlukas.strikingly.com) (MPI EVA, dieter.lukas@eva.mpg.de)'
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    code_folding: hide 
  md_document: 
    toc: true
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  github_document: 
    toc: true
bibliography: /Users/corina/GitHub/grackles/Files/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

***Click [here](https://github.com/corinalogan/grackles/blob/master/README.md) to navigate to the version-tracked reproducible manuscript (.Rmd file)***

###ABSTRACT

In most bird species, females are the dispersing sex because males establish territories and females search for suitable mates (@greenwood1980mating). Because of this, investigations usually do not validate whether this is the case for the particular species of interest. We investigate whether females are the dispersing sex in great-tailed grackles, which have a mating system where the males hold territories and the females choose which territory to place their nest on (@johnson2000male). Results will have bearing on the long-term study on behavioral flexibility and rapid range expansions that is being conducted on this species by allowing us to understand which individuals are likely to experience similar conditions across their lives and which are likely to face new conditions when they become breeders.

###A. STATE OF THE DATA

This preregistration was written in June 2019 after blood was collected and before processing the DNA to obtain the genetic data.

###B. HYPOTHESES

There are sex differences in great-tailed grackle (*Quiscalus mexicanus*) genetic diversity as a result of differential immigration rates. 

**Prediction 1** Average genetic relatedness will be lower among the female DNA samples compared to the male DNA samples. Males are expected to remain close to the area where they hatched, therefore all of the males on the Arizona State University (ASU) campus are expected to come from the same population. In contrast, females are expected to move before their first breeding attempt and therefore the females on campus are likely to come from multiple populations in the broader area because in birds females tend to be the dispersing sex (@greenwood1980mating).

**Prediction 1 alternative 1:** Genetic diversity is higher in males, potentially because males move more than females in this species, or because while fewer males may move between populations, if they do, they move much larger distances and are therefore more likely to bring in divergent genetic variants.

**Prediction 1 alternative 2:** Males and females move equal distances and therefore their genetic diversity is similar. 

**Prediction 1 alternative 3:** The distances that individuals move are smaller or larger than the area of the ASU campus and therefore our samples do not represent the broader population. If the distance moved is within the scale of the campus area, then only very few individuals will have come from outside the campus area and have brought new genetic variants. If the distance moved is broader than the scale of the campus area, then the campus might include multiple populations and therefore males might also be from different populations containing different genetic variants.

**Prediction 2:** There will be a negative correlation between the distance at which two males were trapped and their genetic relatedness. Whereas, for females, we do not expect any association between spatial and genetic distance because in birds females tend to be the dispersing sex (Greenwood 1980). 

**Prediction 2 alternative 1:** There will be no relationship between the distance at which two males were trapped and their genetic relatedness. There are a couple of potential explanations for why this would occur and we are not able to distinguish between them: 1) the distances between trap sites are not large enough to detect a decline in relatedness, 2) males move away from their hatch sites to breed.

###C. METHODS

####**Planned Sample**
	
DNA from 53 great-tailed grackles was obtained from wild adults and juveniles (n=38 adult females, n=15 adult males, juvenile samples were excluded because they had not yet dispersed) caught in Tempe, Arizona, USA. These individuals were either immediately released, or temporarily brought into aviaries for behavioral testing and then released back to the wild.
	
####**Sample size rationale**
	
This was the largest sample size possible by July 2019 when the DNA were analyzed using ddRADseq. 
	
####**Data collection stopping rule**
	
Analyze all blood that was collected by June 2019 when the trapping season stopped.

####**Open data**

When the study is complete, the data will be published in the Knowledge Network for Biocomplexity's data repository.

####**Randomization and counterbalancing** 

No randomization or counterbalancing is involved in this study.

####**Blinding of conditions during analysis**

Experimenters are blind to the sex of the bird when processing samples using ddRADseq (only the alphanumeric bird ID was visible on the tube and no team member has memorized which ID goes with which bird because we give the birds names).

####**ddRADseq**

The DNA will be processed using ddRADseq by Sevchik and Bronwyn Butcher (Cornell University) following methods in @thrasher2018double. These data are post-processed to generate SNP data for relatedness analyses as in @thrasher2018double.

####**Relatedness analyses**

Genetic relatedness between all pairs of individuals is calculated using the package “related” (Pew, Muir, Wang, & Frasier, 2015 https://doi.org/10.1111/1755-0998.12323) in R, following methods in @thrasher2018double. We will use the function 'compareestimators' to assess which relatedness estimator appears to perform the best given the characteristics of our data. 

####**Dependent variable**

Average relatedness between all pairs of individuals

####**Independent variables**

P1: Sex (female, male)

P2: Distance between trap sites (meters)

###D. ANALYSIS PLAN

We do not plan to **exclude** any data, and no data are **missing**. Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam). 

####*Data checking*


####*Ability to detect actual effects*

The number of males in our sample might be too low to detect patterns of isolation-by-distance (change in relatedness within the range of our sampling locations) or to infer the distance of dispersal. Having samples only from a single population might limit our ability to determine frequencies of dispersal, whether just some or the majority of the individuals of one sex or the other disperse. The sample should however be sufficient to obtain a qualitative assessment whether dispersal is more prevalent in either females or in males. 

####*P1: average relatedness and sex*

**Analysis:** 

We will compare the mean and variance in average relatedness among all females to that among all males. Since average relatedness tends to decrease the larger the number of individuals in the sample (regression to the mean), we will perform a permutation analysis to investigate whether the average relatedness among the males in our sample is higher than what would be expected for a random sample of the same number of females or of individuals of either sex. We will repeat 10,000 random draws of 15 individuals either from among the females or from among all individuals, and generate distributions of average relatedness among these samples. We wil assess how the observed average relatedness among the 15 males compares to these random samples.  

####*P2: average relatedness and trap site distance*

**Analysis:** 

To test whether males and females show different isolation-by-distance, we will follow analyses as in (Aguillon et al. 2017 
https://doi.org/10.1371/journal.pgen.1006911). For males and females separately, we will plot genetic relatedness against geographic distance and assess their association using Mantel correlograms. For the analysis, we will initially create 11 distance bins between 0m-500m (the maximum distance between trapping sites). There are roosting trees that are ~50m apart suggesting that dispersal might be occuring at this scale.

###E. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017], SP606267 [2018], and SP639866 [2019])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)
5) University of Cambridge ethical review process (non-regulated use of animals in scientific procedures: zoo4/17 [2017])

###F. AUTHOR CONTRIBUTIONS

**Sevchik:** Hypothesis development, sample processing, data analysis and interpretation, write up, revising/editing.

**Logan:** Hypothesis development, data analysis and interpretation, write up, revising/editing, materials/funding.

**Lukas:** Hypothesis development, data analysis and interpretation, write up, revising/editing, materials/funding.

###G. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

###I. ACKNOWLEDGEMENTS

We thank Bronwyn Butcher and Irby Lovette at the Lab of Ornithology at Cornell University for providing the lab and training for processing the DNA samples using ddRADseq and for post-processing the raw data into a readily analyzable form.

###J. [REFERENCES](MyLibrary.bib)
