---
title: Investigating sex differences in genetic diversity in great-tailed grackles in Tempe, Arizona to infer potential sex biases in dispersal
author: 'August Sevchik (Arizona State University), [Dr. Corina Logan](http://CorinaLogan.com) (Max Planck Institute for Evolutionary Anthropology, MPI EVA, corina_logan@eva.mpg.de), [Dr. Dieter Lukas](http://dieterlukas.strikingly.com) (MPI EVA, dieter.lukas@eva.mpg.de)'
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    code_folding: hide 
  md_document: 
    toc: true
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  github_document: 
    toc: true
bibliography: /Users/corina/GitHub/grackles/Files/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

***Click [here](https://github.com/corinalogan/grackles/blob/master/README.md) to navigate to the version-tracked reproducible manuscript (.Rmd file)***

###ABSTRACT

In most bird species, females are the dispersing sex because males establish territories and females search for suitable mates (@greenwood1980mating). Because of this, investigations usually do not validate whether this is the case for the particular species of interest. We investigate whether females are the dispersing sex in great-tailed grackles, which have a mating system where the males hold territories and the females choose which territory to place their nest on (@johnson2000male). Results will have bearing on the long-term study on behavioral flexibility and rapid range expansions that is being conducted on this species by allowing us to understand which individuals are likely to experience similar conditions across their lives and which are likely to face new conditions when they become breeders.

###A. STATE OF THE DATA

This preregistration was written in June 2019 after blood was collected and before processing the DNA to obtain the genetic data.

###B. HYPOTHESES

There are sex differences in great-tailed grackle (*Quiscalus mexicanus*) genetic diversity as a result of differential immigration rates. 

**Prediction 1** Average genetic relatedness will be lower among the female DNA samples compared to the male DNA samples. Males are expected to remain close to the area where they hatched, therefore all of the males on the Arizona State University (ASU) campus are expected to come from the same population. In contrast, females are expected to move before their first breeding attempt and therefore the females on campus are likely to come from multiple populations in the broader area because in birds females tend to be the dispersing sex (@greenwood1980mating).

**Prediction 1 alternative 1:** Genetic diversity is higher in males, potentially because males move more than females in this species, or because while fewer males may move between populations, if they do, they move much larger distances and are therefore more likely to bring in divergent genetic variants.

**Prediction 1 alternative 2:** Males and females move equal distances and therefore their genetic diversity is similar. 

**Prediction 1 alternative 3:** The distances that individuals move are smaller or larger than the area of the ASU campus and therefore our samples do not represent the broader population. If the distance moved is within the scale of the campus area, then only very few individuals will have come from outside the campus area and have brought new genetic variants. If the distance moved is broader than the scale of the campus area, then the campus might include multiple populations and therefore males might also be from different populations containing different genetic variants.

**Prediction 2:** There will be a negative correlation between the distance at which two males were trapped and their genetic relatedness. Whereas, for females, we do not expect any association between spatial and genetic distance because in birds females tend to be the dispersing sex (Greenwood 1980). 

**Prediction 2 alternative 1:** There will be no relationship between the distance at which two males were trapped and their genetic relatedness. There are a couple of potential explanations for why this would occur and we are not able to distinguish between them: 1) the distances between trap sites are not large enough to detect a decline in relatedness, 2) males move away from their hatch sites to breed.

###C. METHODS

####**Planned Sample**
	
DNA from 53 great-tailed grackles was obtained from wild adults and juveniles (n=38 adult females, n=15 adult males, juvenile samples were excluded because they had not yet dispersed) caught in Tempe, Arizona, USA. These individuals were either immediately released, or temporarily brought into aviaries for behavioral testing and then released back to the wild.
	
####**Sample size rationale**
	
This was the largest sample size possible by July 2019 when the DNA were analyzed using ddRADseq. 
	
####**Data collection stopping rule**
	
Analyze all blood that was collected by June 2019 when the trapping season stopped.

####**Open data**

When the study is complete, the data will be published in the Knowledge Network for Biocomplexity's data repository.

####**Randomization and counterbalancing** 

No randomization or counterbalancing is involved in this study.

####**Blinding of conditions during analysis**

Experimenters were blind to the sex of the bird when processing samples using ddRADseq (only the alphanumeric bird ID was visible on the tube and no team member has memorized which ID goes with which bird because we give the birds names).

####*ddRADseq*

The DNA was processed using ddRADseq by Sevchik and Bronwyn Butcher (Cornell University) following methods in @thrasher2018double. These data were post-processed into an analyzable form as in @thrasher2018double.

####**Dependent variable**

Average relatedness between all pairs of individuals

####**Independent variables**

P1: Sex (female, male)

P2: Distance between trap sites (meters)

###D. ANALYSIS PLAN

We do not plan to **exclude** any data, and no data are **missing**. Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam). 

####*Data checking*

The data will be checked for overdispersion, underdispersion, zero-inflation, and heteroscedasticity with the DHARMa R package [@Hartig2019dharma] following methods by [Hartig](https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html). Note: DHARMa doesn't support MCMCglmm, therefore we will use the closest supported model: glmer from the R package lme4 [@lme4].

####*Ability to detect actual effects*



####*P1: average relatedness and sex*

**Analysis:** 

####*P2: average relatedness and trap site distance*

**Analysis:** 

###E. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017], SP606267 [2018], and SP639866 [2019])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)
5) University of Cambridge ethical review process (non-regulated use of animals in scientific procedures: zoo4/17 [2017])

###F. AUTHOR CONTRIBUTIONS

**Sevchik:** Hypothesis development, sample processing, data analysis and interpretation, write up, revising/editing.

**Logan:** Hypothesis development, data analysis and interpretation, write up, revising/editing, materials/funding.

**Lukas:** Hypothesis development, data analysis and interpretation, write up, revising/editing, materials/funding.

###G. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

###I. ACKNOWLEDGEMENTS

We thank Bronwyn Butcher and Irby Lovette at the Lab of Ornithology at Cornell University for providing the lab and training for processing the DNA samples using ddRADseq and for post-processing the raw data into a readily analyzable form.

###J. [REFERENCES](MyLibrary.bib)