---
title: Does great-tailed grackle space use behavior reflect individual differences in exploration?
author: 'McCune KB (University of California Santa Barbara, UCSB, kelseybmccune@gmail.com), Folsom M (Max Planck Institute for Evolutionary Anthropology, MPI EVA), Ross C (MPI EVA), Bergeron L (UCSB), [Logan CJ](http://CorinaLogan.com) (MPI EVA, corina_logan@eva.mpg.de)'
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    code_folding: hide 
  md_document: 
    toc: true
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  github_document: 
    toc: true
bibliography: /Users/corina/GitHub/grackles/Files/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

***Click [here](https://github.com/corinalogan/grackles/blob/master/README.md) to navigate to the version-tracked reproducible manuscript (.Rmd file)***

###ABSTRACT

This is one of the first studies planned for our long-term research on the role of behavioral flexibility in rapid geographic range expansions. **Project background:** Behavioral flexibility, the ability to change behavior when circumstances change based on learning from previous experience (@mikhalevich_is_2017), is thought to play an important role in a species' ability to successfully adapt to new environments and expand its geographic range (e.g., @lefebvre1997feeding, @griffin2014innovation, @chow2016practice, @sol2000behavioural, @sol2002behavioural, @sol2005big). However, behavioral flexibility is rarely directly tested at the individual level, thus limiting our ability to determine how it relates to other traits, which limits the power of predictions about a species' ability to adapt behavior to new environments. We use great-tailed grackles (Quiscalus mexicanus, a bird species) as a model to investigate this question because they have rapidly expanded their range into North America over the past 140 years (@wehtje2003range, @peer2011invasion) (see an overview of the [5-year project timeline](https://github.com/corinalogan/grackles/blob/master/README.md)). **This investigation:** In this piece of the long-term project we aim to understand whether experimental measures of exploration reflect space use behavior of wild grackles. Exploration, measured in a separate [preregistration] (http://corinalogan.com/Preregistrations/g_exploration.html), is interpreted as an individual’s response to novelty, such as novel environments or novel objects (@reale2007integrating), to gather information that does not satisfy immediate needs (@mettke2002significance). We will measure the space use behavior of wild grackles using radio telemetry to find color-banded birds and record spatial locations over time with GPS. These results will inform whether individual differences in space use behavior are associated with consistent individual differences in exploration, which could be subject to selection, and thus relate to dispersal behaviors influencing this species’ range expansion within populations. Furthermore, if space use behavior correlates with experimental measures of exploration, then space use could be used to inform conservation management strategies (e.g., which individuals are likely to remain in new or restored habitat after a translocation (@may2016predicting) in species where it is not logistically feasible to experimentally measure exploration. Traditional studies of animal movement behavior require spatial and temporal independence of data points for statistical analysis (@swihart1985testing). However, spatial and temporal autocorrelation (where individuals are found in the same locations over time, such that subsequent relocations are predictable based on previous space use) is an intrinsic component of animal behavior and eliminating it can reduce biological relevance (@dray2010exploratory, e.g., animal movement behavior is influenced by the available habitat and resources which are distributed non-randomly across the landscape). Therefore, in addition to using a typical measure of space use, home range size, we propose two new methods for analyzing wild grackle exploratory behaviors. The first will describe individual differences in movement behavior by analyzing autocorrelation of step length (distance between two sequential observations) and turning angle for each individual over time (@pacheco2019nahua), while the second will describe individual differences in spatial preferences by analyzing the repeatability of each individual's occurrence in particular geographic locations.

###A. STATE OF THE DATA

This preregistration uses secondary data: data that are already being collected for other purposes (GPS points in hypothesis 3 and home range sizes in prediction 3 in the [flexibility and foraging](http://corinalogan.com/Preregistrations/g_flexforaging.html) preregistration). This preregistration was written in June 2019, while at the same time increasing the number of GPS points taken per time per bird to provide enough data for the analyses here. 

###B. HYPOTHESES

#### H1: Individual differences in measures of exploration using novel environment and novel object tasks (see separate [preregistration](http://corinalogan.com/Preregistrations/g_exploration.html)) are related to variation in space use (measured via home range size, autocorrelation of step lengths and turning angles, or whether individuals are predictably found in the same locations).

**Prediction 1:** The more exploratory grackles that get closer or make more touches to the novel object and novel environment will be found in a larger expanse (home range size), use less predictable movement patterns (low autocorrelation of step lengths), and occupy a greater variety of spatial locations. This suggests that exploratory individuals are more willing to move into novel areas in the wild.

**Prediction 1 alternative 1:** The more exploratory grackles that get closer or make more touches to the novel object and novel environment will use a smaller amount of space (home range size), use predictable movement patterns (high autocorrelation of step lengths), and consistently occupy the same spatial locations. This suggests that more exploratory individuals may be able to more efficiently use habitat within their home range. For example, in great tits, the slow-exploring phenotype relates to more in-depth investigatory behaviors towards changes in the local environment [@verbeek1994consistent], switching to utilization of different resources in the same area [@van2009personality], and better problem-solving abilities [@cole2011personality]. Therefore it may not be necessary for these individuals to move into new areas for resources such as food or mating opportunities.

**Prediction 1 alternative 2:** Only performance on the novel environment task will correlate positively with space use behavior in the wild. This suggests that perception of, and behavioral interactions with, novel environments (spatial information) differs from that used for novel objects [@mettke2009spatial].

**Prediction 1 alternative 3:** Only performance on the novel object task will correlate positively with space use behavior in the wild. This suggests that, in these urban populations, space use may primarily be driven by grackles searching for novel objects that represent human-provided sources of food.

**Prediction 1 alternative 4:** There will be no correlation between an individual’s proximity or touches to the novel object or novel environment and their space use behavior. This suggests that the experimental measures of exploration either are not relevant enough to how grackles use space in the wild to be able to measure the same trait, or they are independent of space use behavior potentially because the individuals tracked are adults and are already familiar with their home range and surrounding areas and thus do not need to further explore it as if it was novel.

#### H2: Space use behavior will vary among grackles from our three study populations located along different points in the geographic range of this species (core, middle of expansion, and range edge).

**Prediction 2:** Home range sizes will increase, autocorrelation of step lengths and turning angles will decrease (grackle movement behavior will be less predictable), and grackles will use a greater variety of spatial locations as the geographic distance from the original center of the range increases. Specifically, grackles on the edge of the range (likely northern California), will have larger overall home range sizes, exhibit more variety in step lengths and turning angles, and use a greater variety of spatial locations than grackles in the core of the range (Central America). Grackles in the middle of the expanded range will be intermediary in space use (Arizona). This suggests that range expansion likely occurs as a result of individual differences in space use: the individuals on the leading edge of the range expansion show more exploratory movement behavior [@duckworth2007coupling].

**Prediction 2 alternative 1:** Grackles on the edge of the range will have smaller overall home range sizes, high autocorrelation in step length and turning angle, and consistently use the same spatial locations compared to grackles in the middle or core of the current range. This suggests that suitable habitat may be distributed in small patches, that there may be higher predation on grackles that use more space, or that individual grackles specialize on certain novel habitat types that are patchily distributed.

**Prediction 2 alternative 2:** There is no difference across the geographic range in the space use behavior of grackles. This suggests that, on average, all grackles use the same amount of space, or that there is a similar distribution of individual differences in space use in each population. 

###C. METHODS

####**Planned Sample**
	
Great-tailed grackles are caught in the wild and given colored leg bands in unique combinations for individual identification. Some individuals (~60) are brought temporarily into aviaries where we experimentally measure exploration, and then they are released back to the wild. Grackles are individually housed in an aviary (each 244cm long by 122cm wide by 213cm tall) for a maximum of six months where they have ad lib access to water. Grackles are fed Mazuri Small Bird maintenance diet ad lib during non-testing hours (minimum 20h per day), and various other food items (e.g., peanuts, grapes, crackers) during testing (up to 3h per day per bird). Individuals are given three to four days to habituate to the aviaries and then their test battery begins on the fourth or fifth day (birds are usually tested six days per week, therefore if their fourth day in the aviaries occurs on a day off, then they are tested on the fifth day instead). Before release, all grackles are fitted with VHF radio tags (model A2455 by Advanced Telemetry Systems or model BD-2 by Holohil Systems Ltd.) so we can track space use behavior with radio telemetry. Radio tags are attached to the grackles by gluing them to their backs ([@johnson2001great], [@mong2007optimizing]) or by using a leg loop harness (methods as in @rappole1991new) made from sutures (Vicryl undyed 36in sutures, item number D9389 at eSutures.com; 0.5mm diameter, absorbable so they fall off after a few months). 

After release, an experimenter tracks each tagged grackle for approximately 1.5 hours each day, taking a GPS point every 1 minute or every 10 minutes (we will determine the sampling interval that gives the best resolution), regardless of whether the bird moved [@cushman2005elephants]. Researchers are careful to not get too close so that they do not change the bird’s behavior.
	
####**Sample size rationale**
	
We test as many birds as we can in the approximately five years of this study given that the birds are only brought into the aviaries during the non-breeding season (approximately September through March). The minimum sample size for captive subjects will be 57 across the three sites. We catch grackles with a variety of methods, some of which decrease the likelihood of a selection bias for exploratory and bold individuals because grackles cannot see the traps (i.e., mist nets). Once released, we primarily track the space use behavior of the grackles that were in the aviary, but we also collect GPS point locations on all occasions that we see any color-marked bird so we can determine if grackles that were previously in the aviary have different space use behavior after their release.  
	
####**Data collection stopping rule**
	
We will stop collecting GPS location data on tagged birds when home ranges are fully revealed. To determine at what point home ranges have been fully revealed, we will calculate the asymptotic convergence of home range area as in @leo2016home. We will test home range asymptotic convergence for breeding season and non-breeding season movements separately (breeding season: Apr - Aug, non-breeding season: Sep - Mar).

#### **Open materials**

Testing protocols: 
[Exploration protocol](https://docs.google.com/document/d/1sEMc5z2fw6S9C-wVfc2zV331CRPpu3NuA7IhSFUZJpE/edit?usp=sharing) for exploration of new environments and objects, boldness, persistence, and motor diversity.
[Radio tracking protocol](https://docs.google.com/document/d/1jtjgeWJoZ0Q1CfUpV6zdkyQL3p3WfW9KgyLrMNmNMJc/edit?usp=sharing) for attaching radio tags and following birds using radio telemetry.

####**Open data**

When the study is complete, the data will be published in the Knowledge Network for Biocomplexity's data repository.

####**Randomization and counterbalancing** 

There is no randomization in this investigation. The order of the exploration tasks is counterbalanced across birds as in this [separate preregistration] (http://corinalogan.com/Preregistrations/g_exploration.html). The time of day that we collect GPS point locations is counterbalanced within and across birds to account for variation in movement behavior arising from daily circadian rhythms.

####**Blinding of conditions during analysis**

No blinding is involved in this investigation.

####**Dependent variables**

**P1 - P2**

1) Home range size: an estimate calculated with the minimum convex polygon technique, which consists of the smallest polygon to enclose all GPS location points for an individual grackle during its normal activities [@calenge2011home]. Because we are interested in the exploratory movements of grackles, we will not exclude any outlier GPS locations.

2) Autocorrelation of step length: measured as the standard deviation of step lengths

3) Autocorrelation of turning angle: measured as the standard deviation of turning angles

4) Spatial location preference: measured as the repeatability of grackle occurrence in cells of a 5 x 5m grid array across the landscape

One model will be run for each dependent variable

####**Independent variables**

**P1 and P1 alternative 2 - 4**

1) Exploration of novel environment: Latency to approach to 20cm of a novel environment set inside a familiar environment - OR - closest approach distance to a novel environment set inside a familiar environment (choose the variable with the most data)

2) Exploration of novel object: Latency to approach to 20cm of an object (novel or familiar, that does not contain food) in a familiar environment (that contains maintenance diet away from the object) - OR - closest approach distance to an object (novel, familiar) (choose the variable with the most data)

3) Sex: Male or female

4) Condition: Whether the grackle was only ever in the wild or whether it had been temporarily held in aviaries before data collection on space use

**P1 alternative 1**

1) Exploration of novel environment: Latency to approach to 20cm of a novel environment set inside a familiar environment - OR - closest approach distance to a novel environment set inside a familiar environment (choose the variable with the most data)

2) Exploration of novel object: Latency to approach to 20cm of an object (novel or familiar, that does not contain food) in a familiar environment (that contains maintenance diet away from the object) - OR - closest approach distance to an object (novel, familiar) (choose the variable with the most data)

3) Problem-solving performance: Measured as performance on multi-access box tasks described in this [separate preregistration] (http://corinalogan.com/Preregistrations/g_flexmanip.html)

4) Number of different foods eaten and foraging techniques used: Measured during focal follow observations on wild grackles, described in this [separate preregistration] (http://corinalogan.com/Preregistrations/g_flexforaging.html)

3) Sex: Male or female

4) Condition: Whether the grackle was only ever in the wild or whether it had been temporarily held in aviaries before data collection on space use

**P2**

1) Site: Whether the grackle was from our study population located on the edge of the range (most likely Northern California), the population in the center of the original range (Central America), or the population in the center of the current expanded range (Arizona).

2) Sex: Male or female

3) Condition: Whether the grackle was only ever in the wild or whether it had been temporarily held in aviaries before data collection on space use

###D. ANALYSIS PLAN

We do not plan to **exclude** any data and no data are **missing**. Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam). 

We will first verify that the GPS point locations on each bird result in asymptotic convergence as in @leo2016home. To calculate our dependent variables we will use the minimum convex polygon method for estimating home range size (in hectares) using the R packages adehabitatHR [@calenge2006package] and sf [@pebesma2018simple]. Minimum convex polygon is widely used to estimate home range size and only includes points within the areas used by the animal during “normal activities” [@calenge2011home]. Therefore, outlier relocation detections are usually excluded. We are interested in all exploratory movements by grackles, so we will not exclude any outlier relocations.  

From the GPS point locations collected on each individual, we will use a Bayesian model (detailed below) to extract mean and dispersion of step lengths and turning angles for each bird on each daily track. We will determine if the amount of daily autocorrelation is repeatable within individuals across days by analyzing the distribution of the dispersion parameter for step length and turning angle. A narrow distribution would indicate there is low variability (high repeatability) in the daily movement behaviors of the individual.

Moreover, we will determine if grackles show individual differences in consistent use of habitat by overlaying a 5 x 5m grid array across the landscape.  We will then create matrices describing the number of times a grackle was observed in each cell each day. High autocorrelation among each daily matrix indicates a low-exploring individual that remains in the same spatial locations across days.

If movement behavior and spatial location choice is repeatable, then we will use the average value for each individual to model the relationship with performance at our exploration tasks and as a function of study site location. 


#### *Calculating home range size*
```{r h1, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
#Load packages
library(adehabitatHR)
library(sf)

#Point to the correct data file and load it
setwd("~/Documents/Grackle project/Space use")

pts<-read.csv("gtgr_points.csv", header = T)
pts<-pts[,-c(4:6,8:10)]

#Set variables to ensure the models read the data properly
pts$Latitude = as.numeric(as.character(pts$Latitude))
pts$Longitude = as.numeric(as.character(pts$Longitude))
pts$Date = as.character(pts$Date)
pts$Time = as.character(pts$Time)
pts$da.ti = as.POSIXct(paste(pts$Date, pts$Time), format="%Y-%m-%d %H:%M:%S")

### Home range size can only be calculated when a bird has more than 5 relocations
### Count the number of relocations for each bird, then exclude individuals with less than 6
tmp = pts
tmp$count = 1
tmp = aggregate(count ~ Bird.Name, FUN = "sum", data = tmp)
pts.min = merge(pts, tmp, by = "Bird.Name", all = T)
pts.min = pts.min[-which(pts.min$count<6),]
pts.min$Bird.Name = as.factor(as.character(pts.min$Bird.Name))

### Convert dataframe to Spatial file type
p.sf <- st_as_sf(pts.min, coords = c("Longitude", "Latitude"), crs = 4326) 
class(p.sf) 
p.spatial <- as(p.sf, "Spatial")
class(p.spatial)

### Calculate home range in square meters for each individual, excluding outliers
hr = mcp(p.spatial[1], percent = 100, unout = "m2")
plot(hr)
hr = as.data.frame(hr)

### Normalize data for regression model 
hist(log(hr$area)) 

exp <- read.csv("Explore_combined.csv", header = T)
hr_exp = merge(exp, hr, by = "id", all = T)
```

#### *Functions to scaffold the statistical model we will use for analysis of movement behaviors*

These scripts and code are available at https://github.com/ctross/grackleator.

```{r functions, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
### gracklenomial.R
gracklenomial <- function(GrackBins){
model_dat_2=list(
Trips=dim(GrackBins)[1],
Bins=dim(GrackBins)[2],
GrackBins=GrackBins)

model_code_2 <- '
data{
	int Trips;
	int Bins;
	int GrackBins [Trips,Bins];
}

parameters{
 simplex[Bins] A;
 real<lower=0,upper=1> B;
}

model{
 vector[Bins] C;
 A ~ dirichlet(rep_vector(1,Bins));
 B ~ beta(1,1);

 for( i in 2:Trips){
 	C = to_vector(GrackBins[i-1])/sum(GrackBins[i-1]);
	GrackBins[i] ~ multinomial( A*(1-B) + B*C  );
    }
}
'
 m2 <- stan( model_code=model_code_2, data=model_dat_2,refresh=1,chains=1, control = list(adapt_delta = 0.9, max_treedepth = 13))

 return(m2)
}


### gracklepars.R
gracklepars <- function(m1){
MuD_M<-data.frame(Group="Step-Size",Group2="Mean",Group3="Mean",Value=rstan::extract(m1,pars="MuD_M")$MuD_M) # Mean Step-Size over Days
DispD_M<-data.frame(Group="Step-Size",Group2="Mean",Group3="Dispersion",Value=rstan::extract(m1,pars="DispD_M")$DispD_M) # Dispersion in Mean Step-Size over days

MuD_D <-data.frame(Group="Step-Size",Group2="Dispersion",Group3="Mean",Value=rstan::extract(m1,pars="MuD_D")$MuD_D)#  Mean Dispersion in Step-Size 
DispD_D <-data.frame(Group="Step-Size",Group2="Dispersion",Group3="Dispersion",Value=rstan::extract(m1,pars="DispD_D")$DispD_D)# Dispersion in Dispersion in Step-Size 

MuA_M <-data.frame(Group="Heading Change",Group2="Mean",Group3="Mean",Value=rstan::extract(m1,pars="MuA_M")$MuA_M)# Mean Angle Change over Days
DispA_M <-data.frame(Group="Heading Change",Group2="Mean",Group3="Dispersion",Value=rstan::extract(m1,pars="DispA_M")$DispA_M)# Dispersion in Mean Angle Change over days

MuA_D <-data.frame(Group="Heading Change",Group2="Dispersion",Group3="Mean",Value=rstan::extract(m1,pars="MuA_D")$MuA_D)#  Mean Dispersion in Angle Change 
DispA_D <-data.frame(Group="Heading Change",Group2="Dispersion",Group3="Dispersion",Value=rstan::extract(m1,pars="DispA_D")$DispA_D)# Dispersion in Dispersion in Angle Change

df_allx<-rbind(MuD_M,DispD_M,MuD_D,DispD_D,MuA_M,DispA_M,MuA_D,DispA_D)

 ggplot(df_allx, aes(x=Value)) + geom_density(aes(group=Group3, colour=Group3, fill=Group3), alpha=0.3)   + facet_wrap(Group2~Group,scales="free")+
 labs(y="Regression parameters") + theme(strip.text.x = element_text(size=14,face="bold"),axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold")) 
}


### grackletrip.R
grackletrip <- function(m1){
 m1a<-rstan::extract(m1,pars="AlphaAngle")
 sample_eff<-apply(m1a$AlphaAngle,2,quantile,probs=c(0.05,0.5,0.95))
 df_angle<-data.frame(Trip=c(1:dim(m1a$AlphaAngle)[2]),Group="Heading Change",Group2="Mean",
                      LI=sample_eff[1,],
                      Median=sample_eff[2,],
                      HI=sample_eff[3,])
                      
 m1d<-rstan::extract(m1,pars="AlphaDist")
 sample_eff<-apply(m1d$AlphaDist,2,quantile,probs=c(0.05,0.5,0.95))
 df_dist<-data.frame(Trip=c(1:dim(m1d$AlphaDist)[2]),Group="Step-Size",Group2="Mean",
                      LI=sample_eff[1,],
                      Median=sample_eff[2,],
                      HI=sample_eff[3,])
                      
 df_all1<-rbind(df_angle,df_dist)

  m1a<-rstan::extract(m1,pars="DAngle")
 sample_eff<-apply(exp(m1a$DAngle),2,quantile,probs=c(0.05,0.5,0.95))
 df_angle<-data.frame(Trip=c(1:dim(m1a$DAngle)[2]),Group="Heading Change",Group2="Dispersion",
                      LI=sample_eff[1,],
                      Median=sample_eff[2,],
                      HI=sample_eff[3,])
                      
 m1d<-rstan::extract(m1,pars="SDDist")
 sample_eff<-apply(exp(m1d$SDDist),2,quantile,probs=c(0.05,0.5,0.95))
 df_dist<-data.frame(Trip=c(1:dim(m1d$SDDist)[2]),Group="Step-Size",Group2="Dispersion",
                      LI=sample_eff[1,],
                      Median=sample_eff[2,],
                      HI=sample_eff[3,])
                      
 df_all2<-rbind(df_angle,df_dist)

 df_all<-rbind(df_all1,df_all2)

 ggplot(df_all,aes(x=Trip,y=Median))+geom_point()+
 geom_linerange(aes(ymin=LI,ymax=HI))+facet_wrap(Group2~Group,scales="free")+
 labs(y="Regression parameters") + theme(strip.text.x = element_text(size=14,face="bold"),axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold")) + 
 scale_x_continuous(breaks= pretty_breaks())
}


### grackleate.R
grackleate <- function(AlphaDist=2.8,AlphaAngle=0,BetaDist=rep(0,10),BetaAngle=rep(0,10),SDDist=1,DAngle=2,Thresh=50, Lags=10, steps=150, seed=123456, Reps=30,N_Patches=13){

################################################################################# Set up Patches of Food                                                
X_Mushrooms<-vector("list",N_Patches)                                           # Location of prey
Y_Mushrooms<-vector("list",N_Patches)                                           # 

set.seed(seed)                                                                  # Reset Seed

N_PerPatch<-rpois(N_Patches, 200)                                               # Create some prey
for(i in 1:N_Patches){
X_Mushrooms[[i]]<-rnorm(N_PerPatch[i], runif(1,-2500,2500), rpois(1, 350))+100
Y_Mushrooms[[i]]<-rnorm(N_PerPatch[i], runif(1,-2500,2500), rpois(1, 350))+100
}

X_Mushrooms_per_step<-vector("list",steps)                                      # Location of prey
Y_Mushrooms_per_step<-vector("list",steps)                                      #

for(i in 1:steps){
 X_Mushrooms_per_step[[i]]<-X_Mushrooms
 Y_Mushrooms_per_step[[i]]<-Y_Mushrooms
}                                                                 
                                                                                     
################################################################################ Start Model
Store.X <- matrix(NA,ncol=Reps,nrow=steps-1)
Store.Y <- matrix(NA,ncol=Reps,nrow=steps-1)

for(q in 1:Reps){                                                               # Run simulation several times
set.seed(seed)                                                                  # Reset seed

N_PerPatch<-rpois(N_Patches, 200)                                               # Choose number of items per patch

for(i in 1:N_Patches){
X_Mushrooms[[i]]<-rnorm(N_PerPatch[i], runif(1,-2500,2500), rpois(1, 350))+100  # Make some patches of prey
Y_Mushrooms[[i]]<-rnorm(N_PerPatch[i], runif(1,-2500,2500), rpois(1, 350))+100  #
}

X_Mushrooms_per_step<-vector("list",steps)                                      # Location of prey
Y_Mushrooms_per_step<-vector("list",steps)                                      #

for(i in 1:steps){
 X_Mushrooms_per_step[[i]]<-X_Mushrooms
 Y_Mushrooms_per_step[[i]]<-Y_Mushrooms
}

loc.x<-c()                                                                      # Location of forager
loc.y<-c()                                                                      #

speed<-c()                                                                      # Speed or step size
heading<-c()                                                                    # Absolute heading
d.heading<-c()                                                                  # Heading change

Hits<-c()                                                                       # Binary vector of encounters

loc.x[1:Lags]<-rep(0,Lags)                                                      # Intialize vectors
loc.y[1:Lags]<-rep(0,Lags)                                                      #
heading[1:Lags]<-rep(0,Lags)                                                    #
speed[1:Lags]<-rep(0,Lags)                                                      #
Hits[1:Lags]<-rep(0,Lags)                                                       #

plot(loc.x,loc.y,typ="l",ylim=c(-3500,3500),xlim=c(-3500,3500))                 # Plot prey items
 for(i in 1:N_Patches){                                                         #
  points(X_Mushrooms[[i]],Y_Mushrooms[[i]], col="red",pch=".")                  #
   }                                                                            #

set.seed(q*103)

################################################################################ Now model forager movement

for(s in (Lags+1): (steps-1)){                                                  
 X_Mushrooms <- X_Mushrooms_per_step[[s]]
 Y_Mushrooms <- Y_Mushrooms_per_step[[s]]

 PredDist <- AlphaDist;                                                         # First calculate mean prediction conditional on encounters
for(k in 1:Lags){                                                               #    
 PredDist <- PredDist + BetaDist[k]*ifelse(Hits[s-k]>0,1,0);                                  #
       }                                                                        #

 R<- exp(rnorm(1,PredDist,SDDist))                                              # Then simulate a step distance. 
    
 PredAngle <- AlphaAngle;                                                       # Again calculate mean prediction conditional on encounters
for(k in 1:Lags){                                                               #    
 PredAngle <- PredAngle + BetaAngle[k]*ifelse(Hits[s-k]>0,1,0);                              #
        }                                                                       #
                
 Theta<- rbeta(1,inv_logit(PredAngle)*DAngle,                                   # And then simulate a directional change
          (1-inv_logit(PredAngle))*DAngle)*180*ifelse(runif(1,0,1)>0.5,1,-1)    # The 180 shifts from the unit to the maximum absolute distance
                                                                                # The ifelse just chooses a random direction
                                                                             
 heading[s]<-(heading[s-1]+Theta)%%360                                          # Store new heading. Note that the %% is the mod operation to wrap around if needed
 d.heading[s] <- abs(Theta)/180                                                 # Also store just the delta heading
 speed[s] <- R                                                                  # And the speed slash step size
   
 ynew <- R * sin(deg2rad(heading[s]))                                           # Now convert polar to Cartesian, to get the offset for a new x and y pair
 xnew <- R * cos(deg2rad(heading[s]))                                           #
   
 loc.x[s]<-loc.x[s-1]+xnew                                                      # Make the new x and y pair
 loc.y[s]<-loc.y[s-1]+ynew                                                      #

############################################################################### Now check for an encounter 
 Scrap2<-c()
 for(i in 1:N_Patches){                                                         # For each patch
  Scrap<-rep(NA,length(X_Mushrooms[[i]]))
 for(j in 1:length(X_Mushrooms[[i]])){                                          # For each mushroom in patch
 Scrap[j]<- dist(loc.x[s],X_Mushrooms[[i]][j],loc.y[s],Y_Mushrooms[[i]][j]);    # Calculate the distance from the forager to the mushroom
 if(Scrap[j]<Thresh){                                                           # If the forager is closer than the visual radius slash encounter threshold
 points(X_Mushrooms[[i]][j],Y_Mushrooms[[i]][j], col="blue",pch=20)             # Plot a hit

       for(allgone in s: (steps-1)){
       X_Mushrooms_per_step[[allgone]][[i]][j]<-99999                           # If this run is for destructive foraging
       X_Mushrooms_per_step[[allgone]][[i]][j]<-99999                           # disappear food for all future timesteps
      }
 }}
 
 Scrap2[i]<- ifelse(sum(ifelse(Scrap<Thresh,1,0))>0,sum(ifelse(Scrap<Thresh,1,0)),0)      # Check for hits, also can replace sum(ifelse(Scrap<Thresh,1,0)) with 1
 }
 
 Hits[s]<-ifelse(sum(Scrap2)==0,0,sum(Scrap2))                                            # If there is a hit, then set encounters to 1
  
 lines(loc.x,loc.y,ylim=c(-2500,2500),xlim=c(-2500,2500))                                 # Plot updates to the foragers path
}
 Store.X[,q] <- loc.x
 Store.Y[,q] <- loc.y
 }


 return(list(X=Store.X[(Lags+1):length(loc.x),],Y=Store.Y[(Lags+1):length(loc.y),]))
 }


### grackleations.R
grackleations <- function(m2){
df2<-data.frame(Parameter="B",Value=rstan::extract(m2,pars="B")$B) # Mean Step-Size over Days


 ggplot(df2, aes(x=Value)) + geom_density(aes(fill=Parameter), alpha=0.3) +
 theme(strip.text.x = element_text(size=14,face="bold"),axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold")) 
}


### gracklebinner.R
gracklebinner <- function(tracks,nbin=c(15,15),ab_override=NULL){

Trips <- dim(tracks$X)[2]
GrackBins <- matrix(NA,nrow=Trips,ncol=nbin[1]*nbin[2])
bins<-bin2(cbind(c(z$X),c(z$Y)),nbin=nbin)

if(length(dim(ab_override))==0){
for( i in 1:Trips){
GrackBins[i,] <- c(bin2(cbind(z$X[,i],z$Y[,i]),nbin=nbin,ab=bins$ab)$nc)
 }
 } else{
for( i in 1:Trips){
GrackBins[i,] <- c(bin2(cbind(z$X[,i],z$Y[,i]),nbin=nbin,ab=ab_override)$nc)
 }
 }

return(GrackBins)
}


### grackleize.R
grackleize <- function(StoreX,StoreY) {                                          
############################################################ Prepare Data
 Trip <- dim(StoreX)[2]  
 MaxTicks<-dim(StoreX)[1]
    
 Distance <-matrix(NA,ncol=max(Trip),nrow=MaxTicks) 
 Ang <-matrix(NA,ncol=max(Trip),nrow=MaxTicks)
 AngDiff <-matrix(NA,ncol=max(Trip),nrow=MaxTicks)

 N<-c()
 for(j in 1:max(Trip)){
 X <- StoreX[,j]
 Y <- StoreY[,j]
 N[j] <- length(X)
  
 Dist<-c()  
 ang1<-c()
 angdif<-c()
  
 for(i in 2:N[j]){
 Dist[i]<- dist(X[i],X[i-1],Y[i],Y[i-1]);
 ang1[i]<- ang(X[i],X[i-1],Y[i],Y[i-1])
 if(i>3){
 angdif[i]<- ang.dif(ang1[i],ang1[i-1])
         }}    
         
 Distance[1:N[j],j]<-Dist
 Ang[1:N[j],j]<-ang1       
 AngDiff[1:N[j],j]<-angdif
 }

 for(i in 1:MaxTicks){
 for(j in 1:max(Trip)){
  Distance[i,j] <- ifelse(Distance[i,j]==0,
                          runif(1,0,min(Distance[which(Distance != 0)],na.rm=TRUE)),
                          Distance[i,j])
      }}
      

 Distance[is.na(Distance)] <- 999999

 EmptyAngle <-  fitdistr( ScaleBeta(c(na.omit(c(AngDiff)))),start=list(shape1=1,shape2=1),"beta")$estimate

 for(i in 1:MaxTicks){
 for(j in 1:max(Trip)){
  AngDiff[i,j] <- ifelse(is.nan(AngDiff[i,j]), rbeta(1,EmptyAngle[1], EmptyAngle[2])*pi,AngDiff[i,j])
      }}
      
 AngDiff <- ScaleBeta(AngDiff)
 AngDiff[is.na(AngDiff)] <- 999999

 AngDiff <- AngDiff[c(-1,-2,-3),]
 Distance <- Distance[c(-1,-2,-3),]
 N <- N-3
 MaxTicks<-MaxTicks-3


########################################################## Extract data for Stan    
model_dat=list(
Distance=Distance, 
AngDiff=AngDiff,
N=N,
MaxTrip=max(Trip),
MaxTicks=MaxTicks)

################################################################# Fit Stan Model
model_code_1 <- '
data{
int MaxTrip;
int MaxTicks;
int N[MaxTrip];

real Distance[MaxTicks,MaxTrip];
real AngDiff[MaxTicks,MaxTrip];
}

parameters {
 real MuD_M;
 real MuA_M;

 real<lower=0> DispD_M;
 real<lower=0> DispA_M;

 real MuD_D;
 real MuA_D;

 real<lower=0> DispD_D;
 real<lower=0> DispA_D;

 vector[MaxTrip] AlphaDist_Raw;
 vector[MaxTrip] SDDist_Raw; 

 vector[MaxTrip] AlphaAngle_Raw;
 vector[MaxTrip] DAngle_Raw; 
}

transformed parameters{
 vector[MaxTrip] AlphaDist;
 vector[MaxTrip] SDDist; 

 vector[MaxTrip] AlphaAngle;
 vector[MaxTrip] DAngle; 

 AlphaDist = MuD_M + DispD_M*AlphaDist_Raw;
 SDDist = MuD_D + DispD_D*SDDist_Raw;

 AlphaAngle = MuA_M + DispA_M*AlphaAngle_Raw;
 DAngle = MuA_D + DispA_D*DAngle_Raw;  

}

model{
  MuD_M~normal(0,5); 
  MuA_M~normal(0,5); 

  DispD_M~cauchy(0,1); 
  DispA_M~cauchy(0,1); 

  MuD_D~normal(0,5); 
  MuA_D~normal(0,5); 

  DispD_D~cauchy(0,1); 
  DispA_D~cauchy(0,1); 

  AlphaDist_Raw~normal(0,1); 
  SDDist_Raw~normal(0,1); 

  AlphaAngle_Raw~normal(0,1); 
  DAngle_Raw~normal(0,1); 

for(j in 1:MaxTrip){
{

 vector[N[j]] PredAngle;
 vector[N[j]] PredDist;
 vector[N[j]] Dist;
 vector[N[j]] AngleDifference;
  
 for(i in 1:N[j]){
  PredDist[i] = AlphaDist[j]; 
  } 
        
 for(i in 1:N[j]){
  PredAngle[i] = AlphaAngle[j]; 
  }       
              
 for(i in 1:N[j]){            
  Dist[i] = Distance[i,j];              
  AngleDifference[i] = AngDiff[i,j];
  }

 Dist ~ lognormal(PredDist,exp(SDDist[j]));
 AngleDifference ~ beta(inv_logit(PredAngle)*exp(DAngle[j]), (1-inv_logit(PredAngle))*exp(DAngle[j]));
 }}    

}

'

 m1 <- stan( model_code=model_code_1, data=model_dat,refresh=1,chains=1, control = list(adapt_delta = 0.9, max_treedepth = 15))

 return(m1)
}



### setup.R
# Load libraries
	 library(MASS)
	 library(mvtnorm)
	 library(fields)
	 library(ggplot2)
	 library(rethinking)  
	 library(sfsmisc)
	 library(ash)
	 library(reshape)
	 library(maptools)
	 library(gridExtra)
	 library(scales)
	 library(msm)
	 library(maps)
	 library(grid)
	 library(xtable)
	
	# Define functions
	 dist2<-function(a,b){return(sqrt( (b[2]-a[2])^2   + (b[1]-a[1])^2 ))}  
	
	 dist<-function(x2a,x1a,y2a,y1a){return(sqrt((x2a-x1a)^2 + (y2a-y1a)^2))}  
	
	 rad2deg <- function(rad) {(rad * 180) / (pi)}
	
	 deg2rad <- function(deg) {(deg * pi) / (180)}
	
	 ang.dif <- function(x,y) {min((2 * pi) - abs(x - y), abs(x - y))}
	
	 lp_dist <-function(a,b,c){ 
	                           # Return distance between a point and line segment 
	                           t <- -(((a[1]-c[1])*(b[1]-a[1]) + (a[2]-c[2])*(b[2]-a[2])) / ((b[1]-a[1])^2 + (b[2] - a[2])^2 ))
	
	                           if(t >0 & t <1){    
	                            numer <- abs(  (b[2]-a[2])*c[1] -(b[1]-a[1])*c[2] + b[1]*a[2] - b[2]*a[1])
	                            denom <- sqrt( (b[2]-a[2])^2   + (b[1]-a[1])^2 ) 
	                            return(numer/denom)  
	                             }
	
	                           else{
	                            d1 <- dist2(a,c)
	                            d2 <- dist2(b,c)   
	                            return( min(d1,d2))                              
	                             }                                                
	                           }
	
	ScaleBeta <- function(X){
	 a<-0
	 b<-pi
	 y<-X
	 Samp<-50
	
	 y2 <- (y-a)/(b-a)
	 y3<-(y2*(Samp - 1) + 0.5)/Samp
	 y3}
	
	ang<-function(x2,x1,y2,y1){
	  Theta<-seq(0,2*pi,length.out=100)
	  DB<-cbind(cos(Theta),sin(Theta))
	   r <- dist(x2,x1,y2,y1)
	   x0<-(x2-x1)/r
	   y0<-(y2-y1)/r
	
	   (atan2(y0,x0) + pi )
	   }
	
	################################################################# Set parameters
	Thresh <- 50         # Threshold of visual range
	Lags<-10             # Lags at which there are effects of encounters on movement, kinda hardcoded---careful with changes
	Steps<-150           # Length of simulation
	Reps <- 30           # Number of replications
	
	# Adaptive model parameters
	Phi0A <- 3.2
	Psi0A <- 0
	PhiA  <- c(-0.76, -0.65, -0.58, -0.43, -0.29, -0.15, -0.03,0,0,0)
	PsiA <- c(0.5, 0.3, 0.25, 0.2, 0.17, 0.15, 0.13, 0.11, 0.09, 0.08)
	OmegaA <- 0.4
	EtaA <- 2
	
	 # Levy parameters
	Phi0L <- 2.8
	Psi0L <- 0
	PhiL  <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
	PsiL <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
	OmegaL <- 1
	EtaL <- 2
	
	 # Brownian parameters
	Phi0B <- 25
	Psi0B <- 0
	PhiB  <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
	PsiB <- c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
	OmegaB <- 15
	EtaB <- 2
```


#### *Modeling bird movement behaviors: step length, turning angle, spatial location preference*

```{r grackleator, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
#This is an R package for modeling bird movement (available at https://github.com/ctross/grackleator)

#Install by running on R:
library(devtools)
install_github("Ctross/grackleator")

#Some quick examples:

#First, lets look at movement dynamics:

# Load library and attach data
library(grackleator)  

# Simulate tracks from 1 grackle over 30 trips
z <- grackleate(AlphaDist=3.8,AlphaAngle=0,SDDist=1.5,DAngle=2)

# Analyze results of trips for step-size and angle change
m1 <- grackleize(z$X,z$Y)
 
# Plot results over trips
grackletrip(m1)

# Plot bird-specific parameters
gracklepars(m1)

#And now habitat selection:
# Load library and attach data
library(grackleator)  

# Simulate tracks from 1 grackle over 30 trips
z <- grackleate(AlphaDist=3.8,AlphaAngle=0,SDDist=1.5,DAngle=2)

# Bin data on 2D grid
GrackBins<- gracklebinner(z,ab_override=matrix(c(-3000,-3000,3000,3000),nrow=2,ncol=2))

# Model location data
m2 <- gracklenomial(GrackBins)
 
# Plot environmental hotspots
image(matrix(colSums(GrackBins), nrow=15,ncol=15)) # Overall
image(matrix(GrackBins[1,], nrow=15,ncol=15)) # Day 1
image(matrix(GrackBins[2,], nrow=15,ncol=15)) # Day 2
image(matrix(GrackBins[3,], nrow=15,ncol=15)) # Day 3
image(matrix(get_posterior_mean(m2,pars="A"), nrow=15,ncol=15)) # Overall

# Plot bird-specific parameters
grackleations(m2)

# Now create data with temporal correlations
GrackBins2 <- GrackBins
A <- get_posterior_mean(m2,pars="A")
A <- A/sum(A)
B <- 0.8

for(i in 2:30)
GrackBins2[i,] <- rmultinom(1,sum(GrackBins2[i-1,]),A*(1-B) + B*(GrackBins2[i-1,]/sum(GrackBins2[i-1,])))

# Analyze the new data
m3 <- gracklenomial(GrackBins2)

# Plot environmental hotspots
image(matrix(colSums(GrackBins2), nrow=15,ncol=15)) # Overall
image(matrix(GrackBins2[1,], nrow=15,ncol=15)) # Day 1
image(matrix(GrackBins2[2,], nrow=15,ncol=15)) # Day 2
image(matrix(GrackBins2[3,], nrow=15,ncol=15)) # Day 3

image(matrix(get_posterior_mean(m3,pars="A"), nrow=15,ncol=15)) # Post

# Plot bird-specific parameters
grackleations(m3)
```

 

#### *H1: P1 - Exploration measured in captivity relates to space use behavior*

```{r h1, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}

m1 = lm(log(area) ~ ExpObj + ExpEnv + Sex + Condition, data = hr_exp)
hist(m1$resid)
summary(m1)

m2 = lm(log(std_step)) ~ ExpObj + ExpEnv + Sex + Condition, data = hr_exp)
hist(m2$resid)
summary(m2)

m3 = lm(log(std_angle)) ~ ExpObj + ExpEnv + Sex + Condition, data = hr_exp)
hist(H3$resid)
summary(m3)

m4 = lm(log(loc_pref)) ~ ExpObj + ExpEnv + Sex + Condition, data = hr_exp)
hist(H4$resid)
summary(m4)
```

#### *H2: P2 - Space use behaviors vary among populations across the range*

```{r h2, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}

m1 = lm(log(area) ~ Site + Sex + Condition, data = hr_exp)
hist(m1$resid)
summary(m1)

m2 = lm(log(std_step)) ~ Site + Sex + Condition, data = hr_exp)
hist(m2$resid)
summary(m2)

m3 = lm(log(std_angle)) ~ Site + Sex + Condition, data = hr_exp)
hist(m3$resid)
summary(m3)

m4 = lm(log(loc_pref)) ~ Site + Sex + Condition, data = hr_exp)
hist(m4$resid)
summary(m4)
```

###E. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017], SP606267 [2018], and SP639866 [2019])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)

###F. AUTHOR CONTRIBUTIONS

**McCune:** Hypothesis development, data collection, data analysis and interpretation, write up, revising/editing.

**Folsom:** Data collection, revising/editing.

**Ross:** Model development, data analysis and interpretation, revising/editing.

**Bergeron:** Data collection, revising/editing.

**Logan:** Hypothesis development, data interpretation, write up, revising/editing, materials/funding.

###G. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

###I. CONFLICT OF INTEREST DISCLOSURE

We, the authors, declare that we have no financial conflicts of interest with the content of this article. Corina Logan is a Recommender and on the Managing Board at PCI Ecology.

###J. ACKNOWLEDGEMENTS

We thank Ben Trumble for providing us with a wet lab at Arizona State University and Angela Bond for lab support; Melissa Wilson Sayres for sponsoring our affiliations at Arizona State University and lending lab equipment; Kevin Langergraber for serving as local PI on the ASU IACUC; Kristine Johnson for technical advice on great-tailed grackles; Arizona State University School of Life Sciences Department Animal Care and Technologies for providing space for our aviaries and for their excellent support of our daily activities; Julia Cissewski for tirelessly solving problems involving financial transactions and contracts; Richard McElreath for project support; and our research assistants: Aelin Mayer, Nancy Rodriguez, Brianna Thomas, Aldora Messinger, Elysia Mamola, Michael Guillen, Rita Barakat, Adriana Boderash, Olateju Ojekunle, August Sevchik, Justin Huynh, Jennifer Berens, Michael Pickett and Amanda Overholt.

###K [REFERENCES](MyLibrary.bib)
