---
title: Does great-tailed grackle space use behavior reflect individual differences in exploration?
author: 'McCune KB (University of California Santa Barbara, UCSB), Folsom M (Max Planck Institute for Evolutionary Anthropology, MPI EVA), Ross C (MPI EVA), Bergeron L (UCSB), [Logan CJ](http://CorinaLogan.com) (MPI EVA, corina_logan@eva.mpg.de)'
date: '`r Sys.Date()`'
output:
  html_document: 
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
    code_folding: hide 
  md_document: 
    toc: true
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
  github_document: 
    toc: true
bibliography: /Users/corina/GitHub/grackles/Files/MyLibrary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Make code wrap text so it doesn't go off the page when Knitting to PDF
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

***Click [here](https://github.com/corinalogan/grackles/blob/master/README.md) to navigate to the version-tracked reproducible manuscript (.Rmd file)***

###ABSTRACT

This is one of the first studies planned for our long-term research on the role of behavioral flexibility in rapid geographic range expansions. **Project background:** Behavioral flexibility, the ability to change behavior when circumstances change based on learning from previous experience (@mikhalevich_is_2017), is thought to play an important role in a species' ability to successfully adapt to new environments and expand its geographic range (e.g., [@lefebvre1997feeding], [@griffin2014innovation], [@chow2016practice], [@sol2000behavioural], [@sol2002behavioural], [@sol2005big]). However, behavioral flexibility is rarely directly tested at the individual level, thus limiting our ability to determine how it relates to other traits, which limits the power of predictions about a species' ability to adapt behavior to new environments. We use great-tailed grackles (Quiscalus mexicanus, a bird species) as a model to investigate this question because they have rapidly expanded their range into North America over the past 140 years ([@wehtje2003range], [@peer2011invasion]) (see an overview of the [5-year project timeline](https://github.com/corinalogan/grackles/blob/master/README.md)). **This investigation:** In this piece of the long-term project we aim to understand whether experimental measures of exploration reflect space use behavior of wild grackles. Exploration, measured in a separate [preregistration] (http://corinalogan.com/Preregistrations/g_exploration.html), is interpreted as an individual’s response to novelty, such as novel environments or novel objects [@reale2007], to gather information that does not satisfy immediate needs [@mettke2002]. We will measure the space use behavior of wild grackles using radio telemetry to find color-banded birds and record spatial locations over time with GPS. These results will inform whether individual differences in space use behavior are associated with consistent individual differences in exploration, which could be subject to selection, and thus relate to dispersal behaviors influencing this species’ range expansion within populations. Furthermore, if space use behavior correlates with experimental measures of exploration, then space use could be used to inform conservation management strategies (e.g., which individuals are likely to remain in new or restored habitat after a translocation [@May2016]) in species where it is not logistically feasible to experimentally measure exploration. Traditional studies of animal movement behavior require spatial and temporal independence of data points for statistical analysis [@swihart1985].  However, spatial and temporal autocorrelation (where individuals are found in the same locations over time, such that subsequent relocations are predictable based on previous space use) is an intrinsic component of animal behavior and eliminating it can reduce biological relevance ([@Dray2010], e.g., animal movement behavior is influenced by the available habitat and resources which are distributed non-randomly across the landscape). Therefore, in addition to using a typical measure of space use, home range size, we propose to test a novel measure of exploratory behavior in wild grackles using spatial autocorrelation of step lengths (distance between two sequential observations) in our movement data.

###A. STATE OF THE DATA

This preregistration uses secondary data: data that are already being collected for other purposes (GPS points in hypothesis 3 and home range sizes in prediction 3 in the [flexibility and foraging](http://corinalogan.com/Preregistrations/g_flexforaging.html) preregistration). This preregistration was written in June 2019, while at the same time increasing the number of GPS points taken per time per bird to provide enough data for the analyses here. 

###B. HYPOTHESES

#### H1: Individual differences in measures of exploration using novel environment and novel object tasks (see separate [preregistration](http://corinalogan.com/Preregistrations/g_exploration.html)) are related to variation in space use, measured via home range size, and whether individuals are predictably found in the same locations (temporal autocorrelation of step lengths).

**Prediction 1:** The more exploratory grackles that get closer or make more touches to the novel object and novel environment will be found in a larger expanse (home range size), and a greater variety (low autocorrelation of step lengths) of spatial locations. This suggests that exploratory individuals are more willing to move into novel areas in the wild.

**Prediction 1 alternative 1:** The more exploratory grackles that get closer or make more touches to the novel object and novel environment will use a smaller amount of space (home range size), and have less variety in spatial locations. This suggests that more exploratory individuals may be able to more efficiently use habitat within their home range. For example, in great tits, the slow-exploring phenotype relates to more in-depth investigatory behaviors towards changes in the local environment [@Verbeek1994], switching to utilization of different resources in the same area [@Overveld2010], and better problem-solving abilities [@Cole2012]. Therefore it may not be necessary for these individuals to move into new areas for resources such as food or mating opportunities.

**Prediction 1 alternative 2:** Only performance on the novel environment task will correlate positively with space use (home range size) in the wild. This suggests that perception of, and behavioral interactions with, novel environments (spatial information) differs from that used for novel objects [@mettke2009spatial].

**Prediction 1 alternative 3:** Only performance on the novel object task will correlate positively with space use (home range size) in the wild. This suggests that, in these urban populations, space use may primarily be driven by grackles searching for novel objects that represent human-provided sources of food.

**Prediction 1 alternative 4:** There will be no correlation between an individual’s proximity or touches to the novel object or novel environment and their home range size. This suggests that the experimental measures of exploration either are not relevant enough to how grackles use space in the wild to be able to measure the same trait, or they are independent of home range size potentially because the individuals tracked are adults and are already familiar with their home range and surrounding areas and thus do not need to further explore it as if it was novel.

#### H2: Space use behavior will vary among grackles from our three study populations located along different points in the geographic range of this species (core, middle of expansion, and range edge).

**Prediction 2:** Home range sizes will increase and autocorrelation of step lengths will decrease (grackles will use a greater variety of locations) as the geographic distance from the original center of the range increases. Specifically, grackles on the edge of the range (likely northern California), will have larger overall home range sizes, and use a greater variety of spatial locations than grackles in the core of the range (Central America), and grackles in the middle of the expanded range will be intermediary in space use (Arizona). This suggests that range expansion likely occurs as a result of individual differences in space use: the individuals with the larger home range sizes will be on the leading edge of the range expansion [@Duckworth2007].

**Prediction 2 alternative 1:** Grackles on the edge of the range will have smaller overall home range sizes and use a smaller variety of spatial locations than grackles in the middle or core of the current range. This suggests that suitable habitat may be distributed in small patches, that there may be higher predation on grackles that use more space, or that individual grackles specialize on certain novel habitat types that are patchily distributed.

**Prediction 2 alternative 2:** There is no difference across the geographic range in the space use behavior of grackles. This suggests that, on average, all grackles use the same amount of space, or that there is a similar distribution of individual differences in space use at each population. 

###C. METHODS

####**Planned Sample**
	
Great-tailed grackles are caught in the wild and given colored leg bands in unique combinations for individual identification. Some individuals (~57) are brought temporarily into aviaries where we experimentally measure exploration, and then they are released back to the wild. Grackles are individually housed in an aviary (each 244cm long by 122cm wide by 213cm tall) for a maximum of six months where they have ad lib access to water at all times and are fed Mazuri Small Bird maintenance diet ad lib during non-testing hours (minimum 20h per day), and various other food items (e.g., peanuts, grapes, crackers) during testing (up to 3h per day per bird). Individuals are given three to four days to habituate to the aviaries and then their test battery begins on the fourth or fifth day (birds are usually tested six days per week, therefore if their fourth day in the aviaries occurs on a day off, then they are tested on the fifth day instead). Before release, all grackles are fitted with VHF radio tags (model A2455 by Advanced Telemetry Systems or model BD-2 by Holohil Systems Ltd.) so we can track space use behavior with radio telemetry. Radio tags are attached to the grackles by gluing them to their backs ([@Johnson2001], [@Mong2007]) or by using a leg loop harness (methods as in @rappole1991new) made from sutures (Vicryl undyed 36in sutures, item number D9389 at eSutures.com; 0.5mm diameter, absorbable so they fall off after a few months). 

After release, an experimenter tracks each tagged grackle for approximately 1.5 hours each day, taking a GPS point every 1 minute or every 10 minutes (we will determine the sampling interval that gives the best resolution), regardless of whether the bird moved [@Cushman2005].Researchers are careful to not get too close so that they do not change the bird’s behavior.
	
####**Sample size rationale**
	
We test as many birds as we can in the approximately five years of this study given that the birds are only brought into the aviaries during the non-breeding season (approximately September through March). The minimum sample size for captive subjects will be 57 across the three sites. We catch grackles with a variety of methods, some of which decrease the likelihood of a selection bias for exploratory and bold individuals because grackles cannot see the traps (i.e., mist nets). Once released, we primarily track the space use behavior of the grackles that were in the aviary, but we also collect GPS point locations on all occasions that we see any color-marked bird so we can determine if grackles that were previously in the aviary have different space use behavior after their release.  
	
####**Data collection stopping rule**
	
We will stop collecting GPS location data on tagged birds when home ranges are fully revealed. To determine at what point home ranges have been fully revealed, we will calculate the asymptotic convergence of home range area as in @Leo2016. We will test home range asymptotic convergence for breeding season and non-breeding season movements separately (breeding season: Apr - Aug, non-breeding season: Sep - Mar).

#### **Open materials**

Testing protocols: 
[Exploration protocol](https://docs.google.com/document/d/1sEMc5z2fw6S9C-wVfc2zV331CRPpu3NuA7IhSFUZJpE/edit?usp=sharing) for exploration of new environments and objects, boldness, persistence, and motor diversity.
[Radio tracking protocol](https://docs.google.com/document/d/1jtjgeWJoZ0Q1CfUpV6zdkyQL3p3WfW9KgyLrMNmNMJc/edit?usp=sharing) for attaching radio tags and following birds using radio telemetry.

####**Open data**

When the study is complete, the data will be published in the Knowledge Network for Biocomplexity's data repository.

####**Randomization and counterbalancing** 

There is no randomization in this investigation. The order of the exploration tasks is counterbalanced across birds as in this [separate preregistration] (http://corinalogan.com/Preregistrations/g_exploration.html).  The time of day that we collect GPS point locations is counterbalanced within and across birds to account for variation in movement behavior arising from daily circadian rhythms.

####**Blinding of conditions during analysis**

No blinding is involved in this investigation.

####**Dependent variables**

**P1 - P2**

1) Home range size: an estimate calculated with the minimum convex polygon technique, which consists of the smallest polygon to enclose all GPS location points for an individual grackle during its normal activities [@Calange2006].  Because we are interested in the exploratory movements of grackles, we will not exclude any outlier GPS locations.

2) Amount of autocorrelation of successive GPS point locations 

One model will be run for each dependent variable

####**Independent variables**

**P1 and P1 alternative 2 - 4**

1) Exploration of novel environment: Latency to approach to 20cm of a novel environment set inside a familiar environment - OR - closest approach distance to a novel environment set inside a familiar environment (choose the variable with the most data)

2) Exploration of novel object: Latency to land on the table next to an object (novel, familiar) (that does not contain food) in a familiar environment (that contains maintenance diet away from the object) - OR - latency to touch an object (novel, familiar) (choose the variable with the most data)

3) Sex: Male or female

4) Condition: Whether the grackle was only ever in the wild or whether it had been temporarily held in aviaries before data collection on space use

**P1 alternative 1**

1) Exploration of novel environment: Latency to approach to 20cm of a novel environment set inside a familiar environment - OR - closest approach distance to a novel environment set inside a familiar environment (choose the variable with the most data)

2) Exploration of novel object: Latency to land on the table next to an object (novel, familiar) (that does not contain food) in a familiar environment (that contains maintenance diet away from the object) - OR - latency to touch an object (novel, familiar) (choose the variable with the most data)

3) Problem-solving performance: Measured as performance on multi-access box tasks described in this [separate preregistration] (http://corinalogan.com/Preregistrations/g_flexmanip.html)

4) Number of different foods eaten and foraging techniques used: Measured during focal follow observations on wild grackles, described in this [separate preregistration] (http://corinalogan.com/Preregistrations/g_flexforaging.html)

3) Sex: Male or female

4) Condition: Whether the grackle was only ever in the wild or whether it had been temporarily held in aviaries before data collection on space use

**P2**

1) Site: Whether the grackle was from our study population located on the edge of the range (most likely Northern California), the population in the center of the original range (Central America), or the population in the center of the current expanded range (Arizona).

2) Sex: Male or female

3) Condition: Whether the grackle was only ever in the wild or whether it had been temporarily held in aviaries before data collection on space use

###D. ANALYSIS PLAN

We do not plan to **exclude** any data and no data are **missing**. Analyses will be conducted in R (current version `r getRversion()`; @rcoreteam). 

#### *P1 analysis*

We will first verify that the GPS point locations on each bird result in asymptotic convergence as in @Leo2016.  To calculate our dependent variables we will use the minimum convex polygon method for estimating home range size (in hectares) using the R packages adehabitatHR [@calenge2006package] and sf [@pebesma2018simple]. Minimum convex polygon is widely used to estimate home range size and only includes points within the areas used by the animal during “normal activities” [@Burt1943]. Therefore, outlier relocation detections are usually excluded.  We are interested in all exploratory movements by grackles, so we will not exclude any outlier relocations.  

For our second dependent variable, we will quantify the amount of daily autocorrelation in the GPS point locations collected on each individual by comparing Mantel correlograms of a matrix describing distance in space to a matrix of distance in time [@Cushman2005].  

We will determine if the amount of daily autocorrelation is significantly repeatable within individuals across days by dividing our data into 24-hr time periods for each bird and calculating autocorrelation of movement within a day using Mantel correlogram analysis as above. Then we will use Pearson’s product moment correlation to compare space use across the correlograms for each time period. We will use a Monte-Carlo permutation procedure to determine whether the correlation coefficient describing the daily similarity in movement behavior is significantly different from chance [@Cushman2005].   

If movement behavior is repeatable, then we will use the average autocorrelation value for each individual to model the relationship with performance at our exploration tasks and as a function of study site location.  

#### *H1: P1 - Exploration measured in captivity relates to space use behavior*

```{r h1, eval=FALSE, warning=FALSE, results='asis', echo=TRUE, include=TRUE}
#Load packages
library(adehabitatHR)
library(sf)

#Point to the correct data file and load it
setwd("~/Documents/Grackle project/Space use")

pts<-read.csv("gtgr_points.csv", header = T)
pts<-pts[,-c(4:6,8:10)]

#Set variables to ensure the models read the data properly
pts$Latitude = as.numeric(as.character(pts$Latitude))
pts$Longitude = as.numeric(as.character(pts$Longitude))
pts$Date = as.character(pts$Date)
pts$Time = as.character(pts$Time)
pts$da.ti = as.POSIXct(paste(pts$Date, pts$Time), format="%Y-%m-%d %H:%M:%S")

### Home range size can only be calculated when a bird has more than 5 relocations
### Count the number of relocations for each bird, then exclude individuals with less than 6
tmp = pts
tmp$count = 1
tmp = aggregate(count ~ Bird.Name, FUN = "sum", data = tmp)
pts.min = merge(pts, tmp, by = "Bird.Name", all = T)
pts.min = pts.min[-which(pts.min$count<6),]
pts.min$Bird.Name = as.factor(as.character(pts.min$Bird.Name))

### Convert dataframe to Spatial file type
p.sf <- st_as_sf(pts.min, coords = c("Longitude", "Latitude"), crs = 4326) 
class(p.sf) 
p.spatial <- as(p.sf, "Spatial")
class(p.spatial)

### Calculate home range in square meters for each individual, excluding outliers
hr = mcp(p.spatial[1], percent = 100, unout = "m2")
plot(hr)
hr = as.data.frame(hr)

### Normalize data for regression model 
hist(log(hr$area)) 

exp <- read.csv("Explore_combined.csv", header = T)
hr_exp = merge(exp, hr, by = "id", all = T)

H1 = lm(log(area) ~ ExpObj + ExpEnv + Sex + Condition, data = hr_exp)
hist(H1$resid)
summary(H1)
```

###E. ETHICS

This research is carried out in accordance with permits from the:

1) US Fish and Wildlife Service (scientific collecting permit number MB76700A-0,1,2)
2) US Geological Survey Bird Banding Laboratory (federal bird banding permit number 23872)
3) Arizona Game and Fish Department (scientific collecting license number SP594338 [2017], SP606267 [2018], and SP639866 [2019])
4) Institutional Animal Care and Use Committee at Arizona State University (protocol number 17-1594R)

###F. AUTHOR CONTRIBUTIONS

**McCune:** Hypothesis development, data collection, data analysis and interpretation, write up, revising/editing.

**Folsom:** Data collection, revising/editing.

**Ross:** Model development, data analysis and interpretation, revising/editing.

**Bergeron:** Data collection, revising/editing.

**Logan:** Hypothesis development, data interpretation, write up, revising/editing, materials/funding.

###G. FUNDING

This research is funded by the Department of Human Behavior, Ecology and Culture at the Max Planck Institute for Evolutionary Anthropology.

###I. ACKNOWLEDGEMENTS

We thank...

###J. [REFERENCES](MyLibrary.bib)
